---
title: "Coding Lab: `Why code?` and getting situated"
author: "Ari Anisfeld"
date: "Summer 2020"
output: 
  beamer_presentation:
    fig_width: 6
    fig_height: 4
---

<!-- Create an Rmd. install and load packages (tidyverse). Copy and paste some code. Use ?-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
covid_data <- read_csv("~/Downloads/Weekly_counts_of_deaths_by_jurisdiction_and_race_and_Hispanic_origin.csv", col_types = cols(Suppress = col_character())) %>%
    rename(week = `Week Ending Date`,
        time_period = `Time Period`,
         race_ethnicity = `Race/Ethnicity`,
         deaths = `Number of Deaths`,
         diff = `Difference from 2015-2019 to 2020`,
         state = `State Abbreviation`)
```

## Intro to coding lab 

- Why are we here?
- What are we going to do?
- A quick introduction to R and R Studio and the `tidyverse`
 

## 

**Why coding?**

Many public policy jobs and the Harris curriculum rely on programming

- to quickly engage with policy data
- to complete statistical analyses

**Why R?**

- Great data manipulation and visualization suite
- Strong statistical packages (e.g. program evaluation, machine learning)
- Complete programming language with low barriers to entry
- Open source and free

## An example

I wanted to understand racial disparities of Covid-19.


```{r, eval = TRUE, echo = FALSE}
covid_data %>%
  mutate(week = lubridate::mdy(week)) %>%
  filter(race_ethnicity %in% c("Hispanic", "Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Asian")) %>%
  filter(time_period == "2020", MMWRWeek <= 30) %>%
  group_by(race_ethnicity, week ) %>%
  summarize(actual_deaths = sum(deaths, na.rm  = TRUE),
            diff_deaths = sum(diff, na.rm = TRUE),
            expected_deaths = actual_deaths - diff_deaths,
            perc_above_expected = 100 * diff_deaths / expected_deaths) %>%
  ggplot(aes(x = week, color = race_ethnicity)) +
    geom_line(aes(y = perc_above_expected)) + 
  theme_minimal() + 
  labs(y = "Percent of deaths above expected deaths\n(weekly mean 2015-2019 for given week)", 
       x = "", 
       title = "Racial disparities of Covid-19, United States 2020" ,
       subtitle = "Data source: CDC",
       color = "") +
  theme(legend.position = "bottom")
```

## What will we cover?

Foundations:

0. R, RStudio and packages
1. Reading files, and manipulating data with `dplyr`
1. Vectors and data types
1. If statements
1. Analyzing data with groups
1. Basic graph making (summer only)
1. Loops  (in fall)
1. Functions (in fall)

In stats 1 and other courses, you will build off of these lessons:

- extend your capabilities with the functions we teach you
- introduce statistics functions
- introduce new packages etc. based on needs

## Learning philosophy

- We learn coding by experimenting with code.

- Coding is requires a different modality of thinking 

- Coding can be frustrating <!-- I still forget commas and get errors -->

- We develop self-sufficiency by learning where to get help and how to ask for help

- Coding lab is for you. 


## How will we progress?

1. Video lectures:

  - Have R open. Pause regularly. 
  - Focus on main idea first. 
  
2. Practice in labs (most important part):

  - You learn coding by coding.
  - Break up into small groups and work on problems with peer and TA support
  
3. Q and A (live session):

  - Please send me questions ahead of class
  - May include additional practice problems.
  
4. Final project: (see next slide)


## Final project:

  You'll know you're ready for policy school coding, if you can open a data set of interest to you and produce meaningful analysis. For the final project, you will:

  - Pick a data set aligned with your policy interests (or not)
  - Use programming skills to engage with data and make a data visualization 
     showing something you learned from the data.


## Getting help 

- R's `?` documentation is very good, esp. for `tidyverse` code.
- Rstudio has useful cheatsheets for `dplyr` and `ggplot`
  - In the menu bar, select help > cheatsheets
- Get situated with R for Data Science https://r4ds.had.co.nz/
- google and stackoverflow are your friends for idiosyncratic problems

  - googling is its own skill
  - add "in R tidyverse" to your searches for better targeted help


## A quick introduction to R and R Studio and the `tidyverse`

We will

- Discuss what Rstudio is
- Introduce minimal information to get started working with R
- Learn how to install and load packages
- Discuss what the `tidyverse` is


## Getting started with R and R Studio

Please install R and R Studio. These are two distinct things!

We have provided information previously. 

- On the first day, Harris IT will be available for troubleshooting installations.



## What is RStudio?

R Studio is an "integrated developement environment" for R. 

- It provides a console to access R directly.
- A text editor to write R scripts and work with Rmds
- An enviroment and history tab that provide useful information about
what objects you have in your R session
- A help / plots / files / packages etc. section 

## Basic syntax: Variable assignment

We use `<-` for assigning variables in R. 
```{r}
my_number <- 4
my_number
```

## Variable assignment

We can re-assign a variable as we wish. This is useful if we want to try the same math with various different numbers.
```{r eval=FALSE}
my_number <- 2
my_output <- sqrt((12 * my_number) + 1) 
```


## Variable assignment

We assign all sorts of objects to names including data sets and statistical models so that we can refer to them later.

- use names that are meaningful

```{r}
model_fit <- lm(mpg ~ disp +  cyl + hp, mtcars)

summary(model_fit)
```



## Using functions

Functions are procedures that take an input and provide an output.

```{r}
sqrt(4)
median(c(3, 4, 5, 6, 7 ))
```

## Function arguments

Function inputs are called arguments. 

Functions know what the argument is supposed to do based on 

- name
- position

```{r}
f <- function(x, y) {
      2 * x + y
}

f(7, 0)

f(y = 7, x = 0)
```


## Finding help with `?`

```{r, eval=FALSE}
?sum
```

- Description

`sum returns the sum of all the values present in its arguments.`

- Usage  (API)

`sum(..., na.rm = FALSE)`

- Arguments

`...	numeric or complex or logical vectors.`

- Examples (scroll down!)

`sum(1, 2, 3, 4, 5)`


## what are packages?

A package makes a new set of functions available to you. 

Benefits:
- Don't need to code everything from scratch
- Often functions are optimized using C or C++ code to speed up certain steps.

```
Analogy:
- base R comes with screw drivers and hand saws.
- packages give you power tools
```

## installing and loading packages

To use a package we need to:

- install it once from the internet

```{r, eval=FALSE}
install.packages("readxl")  # do this one time directly in console
```

- load it *each time* we restart R

```{r, eval=FALSE}
library(readxl) # add this to your script / Rmd everytime you want to use
read_xlsx("some_data.xls") 
```

- `package::command()` lets you call a function without loading the library

```{r, eval=FALSE}
readxl::read_xlsx("some_data.xls")
```



## common package error

The package 'haven' provides a function to read dta files called `read_dta()`.
What goes wrong here?

```{r, eval=FALSE}
install.packages("haven")
our_data <- read_dta("my_file.dta")

Error in read_dta("my_file.dta") : could not find function "read_dta"
```

## common package error

We need to load the package using `library()`!
```{r, eval=FALSE}
library(haven)
our_data <- read_dta("my_file.dta")
```


## `tidyverse`: set of useful packages

Think of the `tidyverse` packages providing a new dialect for R.

```{r, eval = FALSE}
library(tidyverse)

## -- Attaching packages ----------------------------------
## v ggplot2 3.3.0  v purrr   0.3.4
## v tibble  2.1.3  v dplyr   0.8.5
## v tidyr   1.0.2  v stringr 1.4.0
## v readr   1.3.1  v forcats 0.5.0

## -- Conflicts ------------------------------------------
## x dplyr::filter() masks stats::filter()
## x dplyr::lag()    masks stats::lag()
```


## Recap: Intro to R, RStudio and the `tidyverse`

After going through this video, you should understand how to

- navigate and use Rstudio's features
  - particularly, the console, the text editor and help
- assign objects to names with `<-`
- use functions by providing inputs and learn more with `?`
- `install.packages()` (once) and then load them with `library()` (each time you restart R)


