---
title: "Coding Lab: Reading in data"
author: "Ari Anisfeld"
date: "Summer 2020"
output: 
  beamer_presentation:
    fig_width: 6
    fig_height: 4
---
```{r, setup=TRUE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readxl)
texas_housing_data <- txhousing
```


## So you found some data

Say you find a spreadsheet on the internet and want to start exploring it with R.

## 

Sometimes loading data is as easy as

```{r, eval=FALSE}
texas_housing_data <- read_csv("texas_housing_data.csv")
```

But often you'll need to consider:

- File location
- File type
- Funky formatting

## detour: directory structure

Computer hard drives are organized using a file system. 
In this way, each file has a unique "address" or **file path**. 

- `~/Documents/coding_lab_examples/texas_housing_data.csv` 

The files are stored in folders or directories which are analagous to "zip codes".

- `~/Documents/coding_lab_examples/`


In Windows, file paths start with `C://...`

## detour: working directory

The 'working directory' in an R session is the folder your script knows about. 
If the data you want is in that folder you can refer to it directly.

```{r, eval=FALSE}
fed_data <- 
  read_xlsx("SCE-Public-LM-Quarterly-Microdata.xlsx")
```
 
 `getwd()` shows your current working directory 
<!--point out that if they don't know what a directory structure looks like on their computer (windows) they can see it there -->.

## detour: directory structure

if the data were not in your current working directory you could:

- give the full address: `read_csv("~/Documents/coding_lab_examples/file.csv")`
- give a relative address:
`read_csv("coding_lab_examples/file.csv")`
- change the current working directory: `setwd("~/Documents/coding_lab_examples")`
- move the file to the current working directory: `drag and drop`

## loading data of various formats

We can load data into R with different functions depending on the data format. 
<!--Script: live code read_csv("texas_housing_data.csv")  -->

file type  | package  | function 
-----------|----------|-------------
.csv        | readr    | `read_csv()`
.dta (stata)| haven    | `read_dta()`
.xlsx       | readxl   | `read_xlsx()`

Note: `readr` is loaded with `tidyverse`


## loading data?

While often you can just load the data directly, we often require finesse

```{r}
fed_data <-
  read_xlsx(
    "../data/SCE-Public-LM-Quarterly-Microdata.xlsx")
head(fed_data)
```

<div class="notes">
  open excel workbook. ?read_xlsx
</div>

***
```{r}
fed_data <- 
  read_xlsx(
    "../data/SCE-Public-LM-Quarterly-Microdata.xlsx", 
    sheet = "Data 2013", 
    skip = 1)
head(fed_data)
```


## looking at data: `head()`, `glimpse()` or `View()`

- `head()` and `glimpse()` provide ways to see part of your data.

- `View()` provides a more spreadsheet-like experience.

```{r}
head(texas_housing_data)
```


## getting meta data 

Get number of rows
```{r}
nrow(texas_housing_data) 
```


See column names
```{r}
names(texas_housing_data)
```


## Recap

 - For most file types there's a function of form `read_xxx()` that will get the data into R.
 - Use `getwd()` and `setwd()` to ensure you're in the right directory.
 - When you have funky formatting use `?` to see if R can help you fix the problem on read.
 - R has useful functions like `View()`, `glimpse()`, `head()`, `names()` and `nrow()` to get to know your data.



