---
title: "Coding Lab: If statements and conditionals"
author: "Ari Anisfeld"
date: "Summer 2020"
output: 
  beamer_presentation:
    fig_width: 6
    fig_height: 4
    includes:
      in_header: header_page_n.tex
---

```{r, setup=TRUE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readxl)
options(dplyr.summarise.inform = FALSE) # Suppress summarize info in dplyr >= 1.0
texas_housing_data <- txhousing
```
## Conditional statements (control flow 1)

We often want to our code to do something depending on the context. We start with "if" statements.

```{r eval=FALSE}
if (condition is true) {
  do this
} else {                      
  do this other thing
}
```

In this lesson, we'll

 - review logical operators and comparing vectors
 - introduce `if` and `else` statements
 - introduce vectorized if with `ifelse` in tibbles

## Review: Logical Operators

The logical operators are AND (&), OR (|), and NOT (!). What happens when we use them on booleans?

Let's start with NOT (!).

```{r}
!TRUE
```

```{r}
!FALSE
```

## Review: Logical Operators

Replace the conditional statements 

```{r eval=FALSE}
!(2 > 1)
```

## Review: Logical Operators

Replace the conditional statements 

```{r eval=FALSE}
!(2 > 1)
```

```{r}
!TRUE
```

## What does this produce? {.build}

```{r eval=FALSE}
# NOT (0 does not equal 0)
!(0 != 0)
```

## What does this produce? {.build}


```{r eval=FALSE}
# NOT (0 does not equal 0)
!(0 != 0)
```

```{r}
!FALSE
```

## Review: Logical OR


OR returns `TRUE` if at least one term is `TRUE`.

```{r}
TRUE | FALSE
```

```{r}
FALSE | FALSE
```

Notice that Logical OR has a different meaning than "or" the conjunction  has in common English.

## Review: Logical OR

```{r eval=FALSE}
(5 > 7) | (10 == 10) 
```

## Review: Logical OR

Recall `==` is the logical comparison for if two things are equal.


```{r eval=FALSE}
# 5 is greater than 7 OR 10 equals 10"
(5 > 7) | (10 == 10) 
```
```{r}
FALSE | TRUE
```

## Finally, AND (&)

Returns `TRUE` when both operands are `TRUE`

```{r}
TRUE & FALSE
```


```{r}
TRUE & TRUE
```



## This one is harder... 

```{r eval=FALSE}
!(2 > 6) & (4 > 9 | 3 == 3) 
```


## This one is harder... 

```{r eval=FALSE}
!(2 > 6) & (4 > 9 | 3 == 3) 
```

Break it down:
```{r eval=FALSE}
# Start with the left term
# first
2 > 6
# then
! 2 > 6
```

## This one is harder... 

```{r eval=FALSE}
!(2 > 6) & (4 > 9 | 3 == 3) 
```

Break it down:
```{r eval=TRUE}
# Start with the left term
# first
2 > 6
# then
! 2 > 6
```


## This one is harder... 

```{r eval=FALSE}
!(2 > 6) & (4 > 9 | 3 == 3) 
```

Break it down:
```{r eval=FALSE}
# Now try the right term
# first
4 > 9
# then
3 == 3
# so
(4 > 9 | 3 == 3) 
```

## This one is harder... 

```{r eval=FALSE}
!(2 > 6) & (4 > 9 | 3 == 3) 
```

Break it down:
```{r eval=TRUE}
# Now try the right term
# first
4 > 9
# then
3 == 3
# so
(4 > 9 | 3 == 3) 
```

## This one is harder... {.build}

```{r eval=FALSE}
!(2 > 6) & (4 > 9 | 3 == 3) 
```


```{r}
!(FALSE) & (FALSE | TRUE) 
```


## If statements {.build}

The general syntax of an if statement is as follows:
```{r eval=FALSE}
if (condition is TRUE) {
  do this
}
```

For example:
```{r}
x <- 100

if (x > 0) {
  print("x is positive")
}

```

## If/else statements

Slightly more interesting, the syntax of an if else statement is as follows:

```{r eval=FALSE}
if (condition is TRUE) {
  do this
} else {                      
  do this other thing
}
```

## If/else statements example: 

When working on a project with others, it's sometimes helpful to set 

```{r, eval = FALSE}

if (Sys.info()[["user"]] == "arianisfeld") {
  base_path <- "~/Documents/coding_lab_examples/"
} else {
  base_path <- "~/gdrive/coding_lab_examples/"
} 

data <- read_csv(paste0(base_path, "our_data.csv"))
```


^[Try running `Sys.info()` in your console to understand the code a bit more deeply.]

## multiple tests with `if`, `else if` and `else`


```{r eval=FALSE}
if (condition is TRUE) {
  do this
} else if (second condition is TRUE) {                      
  do this other thing
} else if (third condition is TRUE) {
  do this third thing
} else {
  do a default behavior
}
```

NB: a default behavior with `else` is not necessary.


## multiple tests with `if`, `else if` and `else`
```{r, echo = FALSE}
score <- 10
my_cards <- 12
computers_cards <- 22
```

Here's a cheap version of black jack.
```{r}
score <- 0
my_cards <- sample(2:11, 1) +  sample(2:11, 1)
computers_cards <- sample(2:11, 1) +  sample(2:11, 1)
```

\footnotesize
```{r}
if (my_cards > computers_cards) {
  score <- score + 1
  print("You win")
} else if (my_cards < computers_cards) {
  score <- score - 1
  print("Better luck next time.")
} else {
  print("It's a tie")
}
```

## `if` can take a compound condition

```{r}
if ((my_cards > computers_cards & my_cards <= 21) | 
    computers_cards > 21) {
  score <- score + 1
  print("You win")
} # etc
```

As the statement gets more complex, we're more likely to make errors. 


## `if` is not vectorized and doesn't handle `NA`s
\footnotesize
```{r, eval = FALSE}
if (c(TRUE, FALSE)) { print("if true") }
#> [1] "if true"
#> Warning in if (c(TRUE, FALSE)) {: 
#  the condition has length > 1 and only the
#> first element will be used

if (NA) { print("if true") }
#> Error in if (NA) {: missing value where TRUE/FALSE needed
```

## Vectorized if `ifelse` statements {.build}
\footnotesize
At first blush, `ifelse()` statements look like a quicker way to write an if else statement

```{r}
today <- Sys.Date()
ifelse(today == "2020-11-03", 
       "VOTE TODAY!!", 
       "Don't forget to vote on Nov 3rd.")
```

```{}
ifelse(condition, returns this if TRUE, returns this if FALSE)
```

## What will the following statements return? {.build}

```{r eval=FALSE}
ifelse(TRUE, 1, 2)
ifelse(FALSE, 1, 2)
```

## What will the following statements return? {.build}

```{r eval=TRUE}
ifelse(TRUE, 1, 2)
ifelse(FALSE, 1, 2)
```


## What will the following statements return? {.build}

```{r eval=FALSE}
ifelse(c(TRUE, FALSE, TRUE), 1, 2)
```

## What will the following statements return? {.build}

Unlike `if`, `ifelse` is vectorized! It evaluates item by item.

```{r eval=TRUE}
ifelse(c(TRUE, FALSE, TRUE), 1, 2)
```


## Detour: NAs and missing data

What's going on in this `ifelse()` statement?

```{r echo=TRUE}
ifelse(NA, 1, 2)
```


Unlike `if`, `ifelse` can handle `NA`s and as usual `NA`s are contagious.

## Ifelse statements in dataframes {.build}

Ifelse statements work well in dataframes with the `mutate()` function. 
Let's add a column to the `texas_housing_data` based on a conditional.
\footnotesize
```{r}
texas_housing_data %>% 
  mutate(in_january = ifelse(month == 1, TRUE, FALSE)) %>% 
  select(city, year, month, sales, in_january)
```

## `case_when` statements, supercharged for multiple cases 

If you have a lot of categories, ditch the ifelse statement and use dplyr's `case_when()` function, which allows for multiple conditions, like the `else if`s we saw earlier. 
\footnotesize
```{r}
texas_housing_data %>% 
  mutate(housing_market = 
           case_when(
             median < 100000 ~ "first quartile",
             100000 <= median & median < 123800 ~ "second quartile",
             123800 <= median & median < 150000 ~ "third quartile",
             150000 <= median & median < 350000 ~ "fourth quartile"
           )) %>% 
  select(city, median, housing_market)
```


NB: The conditions in `case_when` can rely on multiple columns. 

## `case_when` statements are a bit "surly"

`case_when` will not do type coercion. 
\footnotesize
```{r, eval = FALSE}
texas_housing_data %>% 
  mutate(housing_market = 
   case_when(
     median < 100000 ~ 1,
     100000 <= median & median < 123800 ~ "second quartile",
     123800 <= median & median < 150000 ~ "third quartile",
     150000 <= median & median < 350000 ~ "fourth quartile"
    )) %>% 
  select(city, median, housing_market)

Error: must be a double vector, not a character vector
Run `rlang::last_error()` to see where the error occurred.
```

Here we try to but doubles and characters in the `housing_market` column, but atomic vectors only have one type!

  - Rather than coerce and provide a warning, the developers decided to make this an error
  - If using `NA` as an output you have to specify `NA` types e.g. `NA_integer_`, `NA_character_`


## Recap: `if` and `ifelse`

Today we learned how to:

- better understand logical operators and conditional statements
- use control flow with `if` and `if`/`else` statements
- use `ifelse()` and `case_when()` statements in conjunction 
with `mutate` to create columns based on conditional statements.




