---
title: "Coding Lab: If Statements and Conditionals"
author: "Ari Anisfeld, Angela Li"
date: "10/15/2019"
output:
  ioslides_presentation
---
```{r, setup=TRUE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readxl)
```

## Lecture Outline

 - Review: Logical operators
 - Comparing vectors
 - If statements
 - Ifelse statements 
  - Detour: NAs and missing data
 - Conditional statements in dataframes

## Review: Logical Operators {.build}

The main logical operators are AND (&), OR (|), and NOT (!). What happens when we use them on booleans?

Let's start with NOT (!).

```{r}
!TRUE
```

```{r}
!FALSE
```

***

## Replace these with conditional statements...{.build}

```{r eval=FALSE}
!(2 > 1)
```

```{r}
!TRUE
```

## What does this produce? {.build}

```{r eval=FALSE}
!(0 != 0)
```

```{r}
!FALSE
```

## How about OR?

```{r}
TRUE | FALSE
```

Only one has to be `TRUE` for the whole statement to return `TRUE`.

## Let's guess again {.build}

Remember that `==` is the logical comparison for if two things are equal.

```{r eval=FALSE}
(5 > 7) | (10 == 10) 
```

```{r}
FALSE | TRUE
```

## Finally, AND (&)

```{r}
TRUE & FALSE
```

Both have to be `TRUE` for the whole statement to return `TRUE`.

## This one is harder... {.build}

```{r eval=FALSE}
!(2 > 6) & (4 > 9 | 3 == 3) 
```
```{r}
!(FALSE) & (FALSE | TRUE) 
```

## Comparing vectors

We've just been comparing numbers, but we can use logical operators to compare vectors. Once again, R *vectorizes* the operation.

```{r}
c(1, 2, 3) > 0
```

```{r}
c(1, 2, 3) > c(0, 2, 4)
```

## choose rows of interest with `filter()`

```{r}
library(tidyverse)
texas_housing_data <- txhousing

texas_housing_data %>%
  filter(year == 2013)
```

***

```{r}
texas_housing_data %>%
  filter(year == 2013, city == "Houston")
```

***

```{r}
texas_housing_data %>%
  filter(year == 2013, city == "Houston" | city == "Galveston")
```

***

```{r}
texas_housing_data %>%
  filter(year == 2013, city %in% c("Houston", "Galveston"))
```



## If statements {.build}

The general syntax of an if statement is as follows:
```{r eval=FALSE}
if (condition is true) {
  do this
}
```

For example:
```{r}
if (2 > 1) {
  print("Math works in R!")
}
```

## If/else statements

Slightly more interesting, the syntax of an ifelse statement is as follows:

```{r eval=FALSE}
if (condition is true) {
  do this
} else {                      
  do this other thing
}
```

## If/else statements, continued

```{r}
today <- "October 15"
my_birthday <- "October 22"
```

```{r}
if (today == my_birthday) {
  print("Happy birthday!")
} else {                      
  print("Not my birthday today.")
}
```

## If/else statements, continued

```{r}
today <- "October 15"
my_birthday <- "October 22"
```

```{r}
if (today == my_birthday) {
  print("Happy birthday!")
} else if (today == "October 31") {                      
  print("It's Halloween!")
} else {
  print("Not a notable day.")
}
```


## Ifelse statements {.build}

If all you need to do is return output based on a condition, `ifelse()` statements are a good choice.

```{r}
ifelse(today == my_birthday, "Happy birthday!", "Not my birthday today.")
```

```{}
ifelse(condition, returns this if TRUE, returns this if FALSE)
```

## What will the following statements return? {.build}

```{r eval=FALSE}
ifelse(TRUE, 1, 2)
ifelse(FALSE, 1, 2)
ifelse(NA, 1, 2)
```

```{r echo=FALSE}
ifelse(TRUE, 1, 2)
ifelse(FALSE, 1, 2)
ifelse(NA, 1, 2)
```

## Detour: NAs and missing data

What's going on in that last `ifelse()` statement?

We have missing data! It's denoted by `NA`.

In this case, the condition needed for the `ifelse()` statement is missing, so our output is also `NA`.

```{r echo=FALSE}
ifelse(NA, 1, 2)
```

## Check for missing data with `is.na()` {.build}

```{r}
is.na(NA)

x <- c(1, 2, NA)
is.na(x)
```

## Remember from last time...`na.rm()`? {.build}

Remove missing values in a vector with `na.rm()`.

```{r}
x <- c(1, 2, NA)
mean(x)
```

```{r}
mean(x, na.rm = TRUE)
```

## Ifelse statements in dataframes {.build}

Ifelse statements work well in dataframes with the `mutate()` dplyr function. Let's add a column to the `txhousing` based on a conditional.

```{r}
texas_housing_data %>% 
  mutate(in_january = ifelse(month == 1, TRUE, FALSE))
```

*** 
Oops, can't see it again - let's use `select()`!

```{r}
texas_housing_data %>% 
  mutate(in_january = ifelse(month == 1, TRUE, FALSE)) %>% 
  select(city, year, month, sales, in_january)
```

## Ifelse statements, supercharged

If you have a lot of categories, ditch the ifelse statement and use dplyr's `case_when()` function, which is essentially a compressed ifelse statement.

```{r}
texas_housing_data %>% 
  mutate(housing_market = 
           case_when(
             median < 100000 ~ "first quartile",
             100000 <= median & median < 123800 ~ "second quartile",
             123800 <= median & median < 150000 ~ "third quartile",
             150000 <= median & median < 350000 ~ "fourth quartile"
           )) 
```

***

```{r}
texas_housing_data %>% 
  mutate(housing_market = 
           case_when(
             median < 100000 ~ "first quartile",
             100000 <= median & median < 123800 ~ "second quartile",
             123800 <= median & median < 150000 ~ "third quartile",
             150000 <= median & median < 350000 ~ "fourth quartile"
           )) %>% 
  select(city, median, housing_market)
```

## Recap

Today we:

- Reviewed logical operators and conditional statements
- Introduced if and if/else statements
- Used ifelse statements in conjunction with `mutate` to create logical columns

Next week: `for` loops

Find the workshop link on the course web page:
http://harris-coding-lab.github.io
