---
title: "Coding Lab: Visualizing data with `ggplot2`"
author: "Ari Anisfeld"
date: "Summer 2020"
output: 
  beamer_presentation:
    fig_width: 3.5
    fig_height: 2
    includes:
      in_header: header_page_n.tex
---

```{r, setup=TRUE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readxl)
library(maps)
options(dplyr.summarise.inform = FALSE)
texas_housing_data <- txhousing
```

## How to use `ggplot`

- How to map data to aesthetics with `aes()` (and what that means)
- How to visualize the mappings with `geom`s 
- How to get more out of your data by using multiple aesthetics
- How to use facets to add dimensionality


*There are whole books on how to use `ggplot`. This is a quick introduction!*



## Understanding `ggplot()`

By itself, `ggplot()` tells R to prepare to make a plot.

```{r}
 texas_annual_sales <- 
  texas_housing_data %>%
  group_by(year) %>%
  summarize(total_volume = sum(volume, na.rm = TRUE))

ggplot(data = texas_annual_sales)
```

## Adding a `mapping`

Adding `mapping = aes()` says how the data will map to "aesthetics".

  - e.g. tell R to make x-axis `year` and y-axis `total_volume`.
  - Each row of the data has (`year`, `total_volume`).  
    - R will map that to the coordinate pair (`x`,`y`) .
    - Look at the data before moving on!

```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) 
```

## Visualizing the mapping with a `geom`

`geom_<name>` tells R what type of visualization to produce.

Here we see points.

- Each row of the data has (`year`, `total_volume`).
- R will map that to the coordinate pair (`x`,`y`).

```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_point()
```

## Visualizing the mapping with a `geom`

Here we see bars.

- Each row of the data has (`year`, `total_volume`).  
- R will map that to the coordinate pair (`x`,`y`) 


```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_col()
```


## Visualizing the mapping with a `geom`

Here we see a line connecting each (`x`,`y`) pair.


```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_line()
```

## Visualizing the mapping with a `geom`
\footnotesize
Here we see a smooth line. R does a statistical transformation!

 - Now R doesn't visualize the mapping (`year`, `total_volume`) to each (`x`,`y`) pair
 - Instead it fits a model to the  (`x`,`y`) and then plots the "smooth" line

```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_smooth()
```




## Visualizing the mapping with a `geom`

We can overlay several `geom`.

```{r, message=FALSE}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_smooth() +
  geom_point()
```


##  Visualizing the mapping with a `geom`


- We saw that we can visualize a relationship between two variables mapping data to `x` and `y`

- The data can be visualized with different geoms that can be composed (`+`) together.

- We can even calculate new variables with statistics and plot those on the fly.


**Next**: Now we'll look at aesthetics that go beyond `x` and `y` axes.

## Using aesthetics to explore data.

We'll use `midwest` data and start with only mapping to `x` and `y`

```{r}
midwest %>%
   ggplot(aes(x = percollege, 
              y = percbelowpoverty)) +
      geom_point()
```

## Using aesthetics to explore data.
\footnotesize

- `color` maps data to the color of points or lines.
  - Each `state` is assigned a color.
  - This works with discrete data and continuous data.

```{r}
midwest %>%
   ggplot(aes(x = percollege, 
              y = percbelowpoverty, 
              color = state)) +
      geom_point()
```

## Using aesthetics to explore data.
\footnotesize
- `shape` maps data to the shape of points.
  - Each `state` is assigned a shape.
  - This works with discrete data only.

```{r}
midwest %>%
   ggplot(aes(x = percollege, 
              y = percbelowpoverty,  
              shape = state)) +
      geom_point()
```

## Using aesthetics to explore data.

\footnotesize
- `alpha` maps data to the transparency of points.
  - Here we map the percentage of people within a known poverty status to `alpha`

```{r}
midwest %>%
   ggplot(aes(x = percollege, 
              y = percbelowpoverty, 
              alpha = poptotal)) +
      geom_point()
```

## Using aesthetics to explore data.
\footnotesize
- `size` maps data to the size of points and width of lines.
  - Here we map the percentage of people within a known poverty status to `size`

```{r}
midwest %>%
   ggplot(aes(x = percollege, 
              y = percbelowpoverty, 
              size = poptotal)) +
      geom_point()
```

## Using aesthetics to explore data.
\footnotesize
We can combine any and all aesthetics, and even map the same variable to multiple aesthetics

```{r, eval = FALSE}
midwest %>%
   ggplot(aes(x = percollege, 
              y = percbelowpoverty, 
              alpha = percpovertyknown,
              size = poptotal,
              color = state))+
      geom_point()
```

## Using aesthetics to explore data.

```{r, echo = FALSE}
midwest %>%
   ggplot(aes(x = percollege, 
              y = percbelowpoverty, 
              alpha = percpovertyknown,
              size = poptotal,
              color = state))+
      geom_point()
```

## Using aesthetics to explore data

Different geoms have specific aesthetics that go with them. 

 - use `?` to see which aesthetics a geom accepts (e.g `?geom_point`) 
    - the bold aesthetics are required.
 - the ggplot cheatsheet shows all the `geom`s with their associated aesthetics
 
## Facets

Facets provide an additional tool to explore multidimentional data 
 
```{r}
midwest %>%
   ggplot(aes(x = log(poptotal), 
              y = percbelowpoverty)) +
      geom_point() +
      facet_wrap(vars(state))
```

## discrete vs continuous data

| aes | discrete | continuous |
| --- | -------- | ----------|
| | limited number of classes | unlimited number of classes | 
| |usually `chr` or `lgl` | numeric |
|`x`, `y`| yes | yes |
| `color`, `fill` | yes | yes |
| `shape`| yes (6 or fewer categories) | no |
| `size`, `alpha` | not advised | yes|
| `facet` | yes | not advised |

Here, discrete and continuous have different meaning than in math

- For `ggplot` meaning is more fluid. 
  - If you do `group_by` with the var and there are fewer than 6 to 10 groups, discrete visualizations can work
  - If your "discrete" data is numeric,  `as.character()` or `as_factor()` to enforce the decision.

## `color` can be continuous

```{r}
midwest %>%
    ggplot(aes(x = percollege, 
               y = percbelowpoverty, 
               color = percpovertyknown)) +
  geom_point()
```

## `shape` does not play well with many categories
\footnotesize
- Will only map to 6 categories, the rest become `NA`.
- We can override this behavior and get up to 25 distinct shapes

```{r, warning = FALSE}
midwest %>%
    ggplot(aes(x = percollege, 
               y = percbelowpoverty, 
               shape = county)) +
  geom_point() + 
  # legend off, otherwise it overwhelms
  theme(legend.position = "none") 
```

## `alpha` and `size` can be misleading with discrete data

```{r}
midwest %>%
    ggplot(aes(x = percollege, 
               y = percbelowpoverty, 
               alpha = state)) +
  geom_point()
```




## Adding vertical lines

```{r}
texas_annual_sales %>%
  ggplot(aes(x = year, y = total_volume)) +
    geom_point() + 
    geom_vline(aes(xintercept = 2007),
                linetype = "dotted")
```

- add horizontal lines with `geom_hline()`
- add any linear fit using `geom_abline()` by providing a slope and intercept.


## Key take aways 

- `ggplot` starts by mapping data to "aesthetics". 
    - e.g. What data shows up on `x` and `y` axes and how `color`, `size` and `shape` appear on the plot.
    - We need to be aware of 'continuous' vs. 'discrete' variables.

- Then, we use `geom`s to create a visualization based on the mapping.
    - Again we need to be aware of 'continuous' vs. 'discrete' variables.
    
- Making quick plots helps us understand data and makes us aware of data issues

**Resources**: R for Data Science chap. 3 (r4ds.had.co.nz); RStudio's ggplot cheatsheet.

## Appendix: Some graphs you made along the way

## lab 0: a map

`geom_path` is like `geom_line`, but connects (`x`, `y`) pairs in the order they appear in the data set.

```{r, eval = FALSE}
storms %>% 
  group_by(name, year) %>% 
  filter(max(category) == 5) %>%
ggplot(aes(x = long, y = lat, color = name)) + 
  geom_path() +
  borders("world") +
  coord_quickmap(xlim = c(-130, -60), ylim = c(20, 50))
```


## lab 0: a map

```{r, echo = FALSE}
storms %>% 
  group_by(name, year) %>% 
  filter(max(category) == 5) %>%
ggplot(aes(x = long, y = lat, color = name)) + 
  geom_path() +
  borders("world") +
  coord_quickmap(xlim = c(-130, -60), ylim = c(20, 50))
```

## lab 1: a line plot

```{r, eval = FALSE, warning=FALSE}
french_data <- 
  wid_data %>% 
  filter(type == "Net personal wealth", 
         country == "France") %>%
  mutate(perc_national_wealth = value * 100)

french_data %>%
  ggplot(aes(y = perc_national_wealth, 
             x = year, 
             color = percentile)) +
  geom_line()
```

## lab 1: a line plot

```{r, echo = FALSE, warning=FALSE}

wid_data_raw <- 
    # You will like have to adjust the file path
    readxl::read_xlsx("../data/world_wealth_inequality.xlsx", 
                      col_names = c("country", "indicator", "percentile", "year", "value")) %>%
    separate(indicator, sep = "\\n", into = c("row_tag", "type", "notes")) 
  
wid_data <- wid_data_raw 
    
french_data <- 
  wid_data %>% 
  filter(type == "Net personal wealth", country == "France") %>%
  mutate(perc_national_wealth = value * 100)

french_data %>%
  ggplot(aes(y = perc_national_wealth, x = year, color = percentile)) +
  geom_line()
```



## lab 2: distributions

- `geom_density()` only requires an `x` asthetic and it calculates the distribution to plot.
- We can set the aesthetics manually, independent of data for nicer graphs.

```{r, eval = FALSE}
    
    chi_sq_samples <-
     tibble(x = c(rchisq(100000, 2), 
                  rchisq(100000, 3), 
                  rchisq(100000, 4)),
            df = rep(c("2", "3", "4"), each = 1e5)) 
    
    chi_sq_samples %>%
      ggplot(aes(x = x, fill = df)) +
      geom_density( alpha = .5) + 
      labs(fill = "df", x = "sample")
```

## lab 2: distributions

```{r,echo = FALSE}
    
    chi_sq_samples <-
     tibble(x = c(rchisq(100000, 1) + rchisq(100000, 1), 
                  rchisq(100000, 3), 
                  rchisq(100000, 4)),
            df = rep(c("2", "3", "4"), each = 1e5)) 
    
    chi_sq_samples %>%
      ggplot(aes(x = x, fill = df)) +
      geom_density( alpha = .5) + 
      labs(fill = "df", x = "sample")
```


## lab 4: grouped bar graphs

- `position = "dodge2"` tells R to put bars next to each other, rather than stacked on top of each other.
- Notice we use `fill` and not `color` because we're "filling" an area.

```{r, eval = FALSE}
    mean_share_per_country %>%
      ggplot(aes(y = country, 
                 x = mean_share, 
                 fill = percentile)) +
      geom_col(position = "dodge2") +
      labs(x = "Mean share of national wealth", 
           y = "", 
           fill = "Wealth\npercentile")
```

## lab 4: grouped bar graphs

```{r, echo = FALSE, warning = FALSE, message = FALSE}

  mean_share_per_country <-
    wid_data %>% 
      filter(percentile %in% c("p99p100", "p90p100")) %>% 
      group_by(country, percentile) %>% 
      summarize(mean_share = mean(value, na.rm = TRUE),
                sd_share = sd(value, na.rm = TRUE))
  
  
  mean_share_per_country_with_time <-
    wid_data %>% 
      filter(country %in% c("China", "India", "USA")) %>%
      filter(percentile %in% c("p99p100", "p90p100")) %>% 
      mutate(time_period = case_when(year < 1960 ~ "1959 and earlier",
                                     year < 1980 ~ "1960 to 1979",
                                     year < 2000 ~ "1980 to 1999",
                                     TRUE ~ "2000 to present")) %>%
      group_by(country, percentile, time_period) %>% 
      summarize(mean_share = mean(value, na.rm = TRUE),
                sd_share = sd(value, na.rm = TRUE))
    
    mean_share_per_country %>%
      mutate(country = case_when(country == "Russian Federation" ~ "Russia", 
                                 country == "United Kingdom" ~ "UK",
                                 country == "South Africa" ~ "S Africa",
                                 TRUE ~ country)) %>%
      ggplot(aes(y = country, x = mean_share, fill = percentile)) +
      geom_col(position = "dodge2") +
      labs(x = "Mean share of national wealth", y = "", fill = "Wealth\npercentile")
```


## lab 4: faceted bar graph

- Notice that we manipulate our data to the right specification before making this graph
- Using `facet_wrap` we get a distinct graph for each time period.

```{r, eval = FALSE}
     mean_share_per_country_with_time %>%
      ggplot(aes(x = country, 
                 y = mean_share, 
                 fill = percentile)) +
        geom_col(position = "dodge2") + 
        facet_wrap(vars(time_period))
```


## lab 4: faceted bar graph

```{r, echo = FALSE, warning = FALSE, message = FALSE}
     mean_share_per_country_with_time %>%
      ggplot(aes(x = country, y = mean_share, fill = percentile)) +
        geom_col(position = "dodge2") + 
        facet_wrap(~time_period)
```

