---
title: "Coding Lab: Visualizing data with `ggplot2`"
author: "Ari Anisfeld"
date: "Summer 2020"
output: 
  beamer_presentation:
    fig_width: 3
    fig_height: 1.5
    includes:
      in_header: header_page_n.tex
---

```{r, setup=TRUE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readxl)
options(dplyr.summarise.inform = FALSE)
texas_housing_data <- txhousing
```

## What will we cover?


- What is the syntax of `ggplot`? What is `aes()`? What are `geom_xxx()`?
- How to use data visualization for exploration?
- How to make data visualization for communication?


## Understanding `ggplot()`

By itself, `ggplot()` tells R to prepare to make a plot.

```{r}
 texas_annual_sales <- 
  texas_housing_data %>%
  group_by(year) %>%
  summarize(total_volume = sum(volume, na.rm = TRUE))

ggplot(data = texas_annual_sales)
```

## Adding a `mapping`

Adding `mapping = aes()` says how the data will map to "aesthetics".

  - e.g. tell R to make x-axis `year` and y-axis `total_volume`.
  - Each row of the data has (`year`, `total_volume`).  
    - R will map that to the coordinate pair (`x`,`y`) .
    - Look at the data before moving on!

```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) 
```

## Visualizing the mapping with a `geom`

`geom_<name>` tells R what type of visualization to produce.

Here we see points.

- Each row of the data has (`year`, `total_volume`).
- R will map that to the coordinate pair (`x`,`y`).

```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_point()
```

## Visualizing the mapping with a `geom`

Here we see bars.

- Each row of the data has (`year`, `total_volume`).  
- R will map that to the coordinate pair (`x`,`y`) 


```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_col()
```


## Visualizing the mapping with a `geom`

Here we see a line connecting each (`x`,`y`) pair.


```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_line()
```

## Visualizing the mapping with a `geom`
\footnotesize
Here we see a smooth line. R does a statistical transformation!

 - Now R doesn't visualize the mapping (`year`, `total_volume`) to each (`x`,`y`) pair
 - Instead it fits a model to the  (`x`,`y`) and then plots the "smooth" line

```{r}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_smooth(se = FALSE)
```




## Visualizing the mapping with a `geom`

We can overlay several `geom`.

```{r, message=FALSE}
ggplot(data = texas_annual_sales,
       mapping = aes(x = year, y = total_volume)) +
  geom_smooth(se = FALSE) +
  geom_point()
```

## Using aesthetics to explore data.
\footnotesize
Now we'll look at aesthetics that go beyond `x` and `y` axes so we can understand our data better.

- `color` maps data to the color of points or lines.
  - Each `state` is assigned a color.
  - This works with discrete data and continuous data.

```{r}
midwest %>%
   ggplot(aes(x = log(poptotal), 
              y = percbelowpoverty, 
              color = state)) +
      geom_point()
```

## Using aesthetics to explore data.
\footnotesize
- `shape` maps data to the shape of points.
  - Each `state` is assigned a shape.
  - This works with discrete data only.

```{r}
midwest %>%
   ggplot(aes(x = log(poptotal), 
              y = percbelowpoverty, 
              shape = state)) +
      geom_point()
```

## Using aesthetics to explore data.
\footnotesize
- `alpha` maps data to the transparency of points.
  - Here we map the percentage of people within a known poverty status to `alpha`^[Using `alpha` for a discrete variable is not advised.]

```{r}
midwest %>%
   ggplot(aes(x = log(poptotal), 
              y = percbelowpoverty, 
              alpha = percpovertyknown)) +
      geom_point()
```

## Using aesthetics to explore data.
\footnotesize
- `size` maps data to the size of points and width of lines.
  - Here we map the percentage of people within a known poverty status to `size`^[Using `size` for a discrete variable is not advised.]

```{r}
midwest %>%
   ggplot(aes(x = log(poptotal), 
              y = percbelowpoverty, 
              size = percpovertyknown)) +
      geom_point()
```

## Using aesthetics to explore data.
\footnotesize
We can combine any and all aesthetics, and even map the same variable to multiple aesthetics

```{r, eval = FALSE}
midwest %>%
   ggplot(aes(x = log(poptotal), 
              y = percbelowpoverty, 
              alpha = percpovertyknown,
              color = state,
              shape = state))+
      geom_point()
```

## Using aesthetics to explore data.
\footnotesize
We can combine any and all aesthetics, and even map the same variable to multiple aesthetics

```{r, echo = FALSE}
midwest %>%
   ggplot(aes(x = log(poptotal), 
              y = percbelowpoverty, 
              alpha = percpovertyknown,
              color = state,
              shape = state))+
      geom_point()
```

## Using aesthetics to explore data

Different geoms have specific aesthetics that go with them. 

 - use `?` to see which aesthetics a geom accepts (e.g `?geom_point`) 
    - the bold aesthetics are required.
 - the ggplot cheatsheet shows all the `geom`s with their associated aesthetics
 
## Facets

Facets provide an additional tool to explore multidimenional data 
 
```{r}
midwest %>%
   ggplot(aes(x = log(poptotal), 
              y = percbelowpoverty)) +
      geom_point() +
      facet_wrap(vars(state))
```
 
## Using aesthetics to explore data

Here's an example of how we can 
 

```{r}
midwest %>%
  mutate(pc = ifelse(perchsd > median(perchsd), "top half", "bottom half")) %>%
  ggplot(aes(x = log(poptotal), 
             y = percbelowpoverty, 
             color = pc, 
             alpha = percpovertyknown / 100)) + 
  geom_point() +
  facet_wrap(~state)
```




## Key take aways 

- `ggplot` starts by mapping data to "aesthetics". 
    - e.g. What data shows up on `x` and `y` axes and how `color`, `size` and `shape` appear on the plot.
    - We need to be aware of 'continuous' vs. 'discrete' variables.
- Then, we use `geom`s to create a visualization based on the mapping.
    - Again we need to be aware of 'continuous' vs. 'discrete' variables.
- Making quick plots helps us understand data and makes aware of data issues
- To communicate effectively with data visualizations, we ...


? What is `aes()`? What are `geom_xxx()`?
- How to use data visualization for exploration?
- How to make data visualization for communication?




```{r}
big_cities <- c("Dallas", "Austin", "San Antonio", "Houston")
texas_housing_cities <- 
  texas_housing_data %>%
   # I'm taking out 2015, since it only has 7 months
   filter(city %in% big_cities, year < 2015) %>%
    group_by(city, year) %>%
  summarize(total_volume = sum(volume, na.rm = TRUE), 
            total_sales = sum(sales, na.rm = TRUE),
            annual_mean_price = total_volume / total_sales
            )
```

## Appendix: Some graphs you made along the way

```{r}
storms %>% 
  group_by(name, year) %>% 
  filter(max(category) == 5) %>%
ggplot(aes(x = long, y = lat, color = name)) + 
  geom_path() +
  borders("world") +
  coord_quickmap(xlim = c(-130, -60), ylim = c(20, 50))
```

