---
title: "Coding Lab: iteration and loops"
author: "Ari Anisfeld"
date: "Fall 2020"
output: 
  beamer_presentation:
    fig_width: 3.5
    fig_height: 2
    includes:
      in_header: header_page_n.tex
---
```{r, setup=TRUE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readxl)
```

## Iteration and for-loops (Control flow II)

We use for-loops to repeat a task over many different inputs or to repeat a simulation process several times.

- How to write for-loops
- When to use a for-loop vs vectorized code

```{r}
for(value in c(1, 2, 3, 4, 5)) {
  print(value)
}
```


## Simple for-loop

```{r}
for (x in c(3, 6, 9)) {
  print(x)
}
```

## Simple for-loop: what is going on?


```{r, eval = FALSE}
for (x in c(3, 6, 9)) {
  print(x)
}
```

Our for-loop is equivalent to the following code. For each value in `c(3,6,9)`, we assign the value to `x` and the do the action between the curly brackets in order.

```{r, eval = FALSE}
x <- 3
print(x)
x <- 6
print(x)
x <- 9
print(x)
```

## For loops

The general structure of a for loop is as follows:

```{r eval=FALSE}
for (value in list_of_values) {
  do something (based on value)
}

for (index in list_of_indices) {
  do something (based on index)
}
```


## Example: find sample means

Suppose we want to find the means of increasingly large samples.

```{r}
mean1 <- mean(rnorm(5))
mean2 <- mean(rnorm(10))
mean3 <- mean(rnorm(15))
mean4 <- mean(rnorm(20))
mean5 <- mean(rnorm(25000))

means <- c(mean1, mean2, mean3, mean4, mean5)

means
```


## Example: find sample means

Let's avoid repeating code with a `for` loop. 

```{r}
sample_sizes <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(sample_sizes))

for (i in seq_along(sample_sizes)) {
  sample_means[[i]] <- mean(rnorm(sample_sizes[[i]]))
}

sample_means
```

In the following slides we'll explain each step.

## Finding sample means, broken down 

Assign initial variables **before** starting the for loop.

```{r eval=FALSE}
# determine what to loop over
sample_sizes <- c(5, 10, 15, 20, 25000)

# pre-allocate space to store output
sample_means <- rep(0, length(sample_sizes))
```

To start: 

1. create a vector of the `sample_sizes` we want to use
2. create a vector to store the output 

## What does `sample_means` currently look like?


```{r}
sample_means <- rep(0, length(sample_sizes))
sample_means
```

**Why do this?** It makes the code more efficient. An alternative is to build up an object as you go, but this requires copying the data over and over again.


## Alternative ways to preallocate space

```{r, eval=FALSE}
sample_means <- vector("double", length = 5)
sample_means <- double(5)
```

Each data type has a comparable function e.g. `logical()`, `integer()`, `character()`.


To hold data of different types, we'll use lists.^[Lists are vectors that are not "atomic".]

```{r, eval=FALSE}
data_list <- vector("list", length = 5)
```

## Adding data to a vector, broken down 

Determine what sequence to loop over. 

```{r eval=FALSE}
for (i in 1:length(sample_sizes)) {

}
```


## A helper function `seq_along()`

`seq_along(x)` is synonymous to `1:length(x)` 

where `x` is a vector.

**Example**
```{r}
vec <- c("x", "y", "z")
1:length(vec)
seq_along(vec)
```

***
```{r}
sample_sizes <- c(5, 10, 15, 20, 25000)
seq_along(sample_sizes)
```

(What if `sample_sizes` is accidentally a 0-length vector? See [what happens](https://r4ds.had.co.nz/iteration.html#for-loops) in R for Data Science.)


## Adding data to a vector, broken down 

```{r eval=FALSE}
sample_sizes <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(sample_sizes))

for (i in seq_along(sample_sizes)) {

}
```

Use `seq_along()` to be safe! 


## Adding data to a vector, broken down {.build}

```{r}
sample_sizes <- c(5, 10, 15, 20, 25000)
sample_means <- numeric(length(sample_sizes))

for (i in seq_along(sample_sizes)) {
  
  sample_means[[i]] <-  mean(rnorm(sample_sizes[[i]]))
  
}

sample_means
```

Save the mean of the sample to the ith place of the `sample_means` vector.


## Common error.

This code falls, because we do not store the output in `sample_means` in the for loop! (Compare to previous slide).

```{r}
sample_sizes <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(sample_sizes))

for (i in seq_along(sample_sizes)) {
  mean(rnorm(sample_sizes[[i]]))
}

sample_means
```

Right now, we're calculating the mean, but it's not being saved anywhere.

## Another example

You get data stored in split over several `csv` files. 

We can read the data into R and store it store it as a single data set.

```{r, eval = FALSE}
setwd("../data/loops")

file_1 <- read_csv("data_1999.csv")
file_2 <- read_csv("data_2000.csv")
...
file_22 <- read_csv("data_2020.csv")

data <- bind_rows(file_1, file_2, ..., file_22)
```

## Aside: how to make the data

The data used for this exercise is fake data which I made with a for-loop.
Run the code below (*choose your own working directory*) to follow along.

```{r, eval=FALSE}
setwd('../data/loops')

file_list <- paste0("data_", 1999:2020, ".csv")

for (file in file_list) {
  data <- 
    tibble(id = 1:100,
           employed = sample(c(0, 1, 1, 1), 
                             100, replace = TRUE),
             happy = sample(c(0,1), 
                            100, replace = TRUE))
  
  write_csv(data, file)
}
```



## Aside: bind_rows()?

`bind_rows()` stacks two dataframe, or combines two vectors into a dataframe:

```{r}
df_1 <- tibble(col1 = 1, col2 = "A")
df_2 <- tibble(col1 = 2:3, col2 = c("B", "C"))

bind_rows(df_1, df_2)
```

## Aside: `list.files()`

`?list.files()`:

These functions produce a character vector of the names of files ... in the named directory.

  - `pattern` ensures we only take the csv files. 
  - It uses *regular expressions* where `*` in `*.csv$` matches any string and `.csv` ensures the string ends in `csv`.

```{r}
list.files("../data/loops", pattern = "*.csv$")
```

## Let's use a loop to read in the data

```{r, eval=FALSE}
file_names <- list.files(pattern = "*.csv$")

output <- vector("list", length(file_names))

for (i in seq_along(file_names)) {
  output[[i]] <- read_csv(file_names[[i]]) %>%
                    mutate(year = str_extract(file_names[[i]], "[0-9]{4}"))
}

data <- bind_rows(output)
View(data)
```


## Let's use a loop to read in the data...an alternative

```{r, message = FALSE}
setwd('../data/loops')

# by default, reads files in working directory
file_list <- list.files(pattern = "*.csv$")

out <- tibble()

for (file in file_list) {
  temp <- read_csv(file)
  
  out <- bind_rows(out, temp)
}

nrow(out)
```



## Review: Vectorized operations

When possible, take advantage of the fact that R is vectorized.

```{r}
a <- 7:11
b <- 8:12
out <- rep(0L, 5)

for (i in seq_along(a)) {
  out[[i]] <- a[[i]] + b[[i]]
}

out
```

This is a bad example of a for loop!

## The better alternative: vectorized addition

```{r}
a <- 7:11
b <- 8:12
out <- a + b

out
```

Use vectorized operations and tidyverse functions like `mutate()` when you can.



## Key points: iteration

- Iteration is useful when we are repeatedly calling the same block of code or function while changing one (or two) inputs. 

- If you can, use vectorized operations.

- Otherwise, for loops work for iteration

    - Clearly define what you will iterate over (values or indicies)
    - Preallocate space for your output (if you can)
    - The body of the for-loop has parametrized code based on thing your iterating over
    - Debug as you code by testing your understanding of what the for-loop should be doing (e.g. using `print()`)
    
    
    
**Further study**: Many R coders prefer the `map()` family functions from `purrr` or base R `apply` family. See [iteration](https://r4ds.had.co.nz/iteration.html) in R for Data Science 