---
title: "Functions and Coding lab in Review"
author: "Ari Anisfeld"
date: "10/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```



## Loops in review

Re-write the loop as vectorized code.

1. Set-up: You want to graph the function $y = 4e^{-4x}$

```{r}
x <- 1:20 / 20 
y <- rep(0, length(x))

for (i in seq_along(x)) {
  # exp() gives use e to a power.
  e_power <- exp(-4*x[[i]])
  y[[i]] <- 4 * e_power
}

qplot(x, y, geom="line")
```

2.  Set-up: you got some data where years are not numeric and only provide the decade, but you want numeric years in order.

```{r}
years <- c("85", "90", "95", "00", "05", "10")
full_years <- c()

for (yr in years) {
  yr <- as.numeric(yr)
  
  if (yr >= 85) {
    temp <- 1900 + yr
  } else {
    temp <- 2000 + yr
  }
  
  full_years <- c(full_years, temp)
  
}

full_years
```

3. Use case: you have a vector of weekly salaries and want to calculate the annual salary

```{r}
set.seed(100)
weekly_salary <- runif(52, min = 400, max = 800)

annual_salary <- 0

for (wage in weekly_salary) {
  annual_salary <- annual_salary + wage
}

annual_salary
```



## Vectorized code in a data context.

Take your code from the previous exercises and incorporate them into a data context.

1. Use mutate to create a column called $y = 4e^{-4x}$, then plot your line.

```{r, eval = FALSE}
data <- tibble(x = 1:20 / 20)



____ %>% # your adjusted data
  ggplot(aes(x=x, y=y)) +
  geom_line()
```


2.  Use mutate to recreate the column years with numeric values.

```{r, eval = FALSE}
econ_data <- tibble(years = c("85", "90", "95", "00", "05", "10"),
               gdp = c(1490, 2560, 4825, 9370, 14900, 16700))


____ %>% # your adjusted data
  ggplot(aes(x=years, y=gdp)) +
  geom_line()
  
```

3. We use group_by and summarize to show the annual salaries of two people in our fake data set.

```{r, eval = FALSE}
# We make fake_data with two people in it to justify using group_by(). 
fake_data <- tibble(weekly_salary = c(runif(52, min = 400, max = 800),
                               runif(52, min = 500, max = 900)),
                    person = c(rep(c(1, 2), each = 52)))
 
fake_data %>%
  group_by() %>%
  summarize()

```



## Functions

# Example 1

1. Use case: We have a survey where the missing values in several columns are indicated by -999. We want to replace -999 with `NA`.
There are many columns, so we want a function to do it.

To fix ideas consider the fake survey data below.
```{r}
fake_survey_data <- tibble(id = 1:5,
                           age = c(30, -999, 40, 43, -999),
                           income = c(6e5, -999, -999, 1e6, 1e6))

fake_survey_data
```


Let's simplify our problem even further. Here are minimial input and output.

```{r}
## input: a vector and missing value indicator
example_input <- c(30, -999, 40, 43) 
## output: a vector with missing values
desired_output <- c(30, NA, 40, 43)
```

Let's write the function `set_nas` togther:
```{r, eval = FALSE}

```

Now let's test it.
```{r, eval = FALSE}
set_nas(example_input)
```


Finally, what should we do if we find out there are several different numbers used to show missing values e.g. there are indicators for "refused to answer", "unknown" and simply "missing". 

Adjust `set_nas()` above so that we remove all the missing value indicators below.

```{r, eval = FALSE}
fake_survey_data <- tibble(id = 1:6,
                           age = c(30, -999, 40, 43, -999, -888),
                           income = c(6e5, -999, -999, 1e6, 1e6, -777))

fake_survey_data %>%
  mutate(age = set_nas(age, c(-999, -888)),
         income = set_nas(age, c(-999, -777)))
```


## Example 2

2. You want to create a pitch deck that shows this graph for various cities.

```{r, echo = FALSE}
txhousing %>%
  filter(city == "Austin") %>%
  group_by(city, year) %>%
  summarize(sales = sum(sales)) %>%
  ggplot(aes(x = year, y = sales)) + 
  geom_line() +
  labs(title = "Austin's annual sales volume")
```

**Think:** Can I parameterize this code so that it still works?

```{r, echo = FALSE}

```

Now wrap it with a function! called `sales_trends`

```{r}

```

Test your code! We should see a plot with Houston's data.

```{r, eval = FALSE}
sales_trends("Houston")
```



## you try

Use case: You're analyzing two surveys together. In one survey a number of variables are *top coded*.

Survey 1: Cut off variables at `$60000`

|id|consumption | income| 
|--|--|---|
| 1 | 60000 | 60000 |
| 2 | 50400 | 60000 |
| 3 | 14000 |24000 |
| ... | ... | ...|


Survey 2: (different people) not top coded

|id|consumption | income| 
|--|--|---|
| "b1" | 14000 | 50000 |
| "b2" | 40400 | 70000 |
| "b3" | 95000 | 120000 |
| ... | ... | ...|

To compare the two surveys, we may want to "censor" the large values in survey 2. 

- what is the input? 
- what output do we expect?
- what is a minimal example? 

Here's an example of minimal inputs and output

input: 

- `vector <- c(1, 2, 30, 2)`
- `threshold <- 5`

output:
- `c(1, 2, 5, 2)`

Write a function called `censor` to top code the data!

```{r}

```


You wrote `censor`! But then you realize some things are thresholded from below!

e.g.

input: 
  
- vector <- c(1, 2, 30, 2) 
- threshold <- 5

output:
- c(5, 5, 30, 5)

## Extending funtionality

- Add an input argument called `above` that is logical (TRUE or FALSE)

Write a new version of `censor`!

```{r, eval = FALSE}


# tests
censor(c(1, 2, 30, 2), 5)
censor(c(1, 2, 30, 2), 5, above = FALSE)
```


**Hint**

- Use an `if()` statement with `above` to determine which direction to censor.

```{r, eval = FALSE}
# This is a skelton for the **body** of your function!
if(above==TRUE){
  # censore high values
  ...
} else {
  # censore low values
  ...
}
```




## Bonus material -- passing column names to functions
Extra slick: can we parameterize the `y` variable.

```{r, eval = FALSE}
trends <- function(city_name, var, var_description){
txhousing %>%
  filter(city == city_name) %>%
  group_by(city, year) %>%
    summarize(y = sum({{ var }})) %>%
  ggplot(aes(x = year, y = y)) + 
  geom_line() +
  labs(title = str_c(city_name, 
                     "'s ", 
                     var_description))
}

trends("Austin", listings, "annual listings")
```

What is ``{{ }}``? 

``{{ }}`` "forwards" the name `var`  from the tibble's environment (where it's a column name) to your global environment (where it's undefined)

- The ideas here are somewhat advanced for coding camp 



