---
title: "qa material"
author: "Ari Anisfeld"
date: "9/3/2020"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)


df_raw <- read_csv("../data/Coding Camp Feedback (Responses) - Form Responses 1.csv")

df <-
  df_raw %>%
  mutate(group = str_extract(`Which TA / Section group are you in?`, "C[12][AB]"),
         cycle = ifelse(str_detect(group, "A"), "A", "B"),
         ta = str_extract(`Which TA / Section group are you in?`, "[A-Z][A-Za-z][a-z]*"),
         ta = str_to_sentence(ta),
         exp = `Describe your past experience (Check ALL boxes that apply).`,
         exp_r = case_when(str_detect(exp, "starting coding") ~ 0L,
                         str_detect(exp, "beginner in R") ~ 1L,
                         str_detect(exp, "intermediate in R") ~ 2L,
                         str_detect(exp, "proficient in R") ~ 3L,
                         TRUE ~ NA_integer_),
         exp_other = case_when(str_detect(exp, "starting coding") ~ 0L,
                               str_detect(exp, "beginner in a") ~ 1L,
                               str_detect(exp, "intermediate .+in a") ~ 2L,
                               str_detect(exp, "proficient in a") ~ 3L,
                               TRUE ~ NA_integer_),
         exp = exp_r > 0 | is.na(exp_r), TRUE, FALSE,
         exp = ! exp,
         pace = `How is the current pace of the class?`,
         comfort = `Do you feel comfortable asking questions?`,
         recs_more_qs = `Do you have any recommendations for how we can make it easier/more comfortable for students to ask questions?`,
         likes = `Are there any aspects of your experience so far that you particularly like (this helps us know what we should continue doing/have more of)?`,
         suggestions = `Any other comments or suggestions?`
            ) %>%
  select(cycle, group, ta, starts_with("exp"), pace, comfort, recs_more_qs, likes, suggestions)

feedback_data <- df %>% distinct()



library(haven)
mrc_data <- read_dta("../data/mrc_table10.dta")

wid_data_raw <- 
        # You will like have to adjust the file path
        readxl::read_xlsx("../data/world_wealth_inequality.xlsx", 
                          col_names = c("country", "indicator", "percentile", "year", "value")) %>%
        separate(indicator, sep = "\\n", into = c("row_tag", "type", "notes")) 
  
    wid_data <- wid_data_raw %>% 
                  select(-row_tag) %>%
                  select(-notes, everything()) %>% 
                  mutate(type = ifelse(str_detect(type, "Net personal wealth"), 
                                       "Net personal wealth", type)) %>%
                  filter(type == "Net personal wealth")

```

## Key points: class 0

- Rstudio has a console to access R and a text editor to write code for reproducible projects.
  - Analogy: R is to RStudio as Tony Stark is to Ironman's suit
- R extensible through packages
  -  use `install.packages("")` once and then `library()` each session
- Use `<-` to assign *any* object to a name
- functions take inputs and return outputs
  - input "understood" based on position or name
  - find out more about functions with `?` (e.g. `?filter`)
  
**Up next:** Watch videos for lab 1: reading data into R and manipulating it with `dplyr` verbs.
  
## Questions from TAs / Piazza

- How do you assign a name to data? (`<-`)
- What is the difference between console and text editor section of R markdown?
- What is `%>%`?
- How do  I read errors?
- What's the deal with backticks and quotes?

- Please add additional questions to the chat.


# Reading files and `dplyr`

## Key points: class 1: Reading files

- Tabular data is stored in a lot of different formats
  - e.g. `.csv`, `.xlsx`, `.dta`
- Read tabular data of a given type with the proper function
  - e.g. for `csv`s we have `read_csv()`
  - if you get a new type, google "How to read xxx files into R tidyverse"
- We need to be aware of the file path and can `setwd()`
- We know there are useful tools built into the `read_xxx()` functions
  - though we just scratched the surface

## Questions from piazza / ta sessions.

- What do the errors mean? Can you write code to get the same errors on your machine?

```{r, eval = FALSE, echo = TRUE}
setwd("~/Docutents")
Error in setwd("~/Docutents") : 
  cannot change working directory
```

```{r,eval = FALSE}
setwd(~/Documents/)
Error: unexpected '/' in "setwd(~/"
```

```{r,eval = FALSE}
 # Finally I got it to work
setwd("~/Documents")
```

## Questions from piazza 

- @35 a funky data issue in excel files on some Windows machines

- Why does `read_dta("some_name.RData")` fail?

## Key points: class 1: Manipulating data with `dplyr()`

- Choose columns with `select()`
- Choose rows based on a match criteria with `filter()`
  - We were introduced to comparison operators like  `==` and `%in%` (more in class 3)
- Make new columns with `mutate()`
  - We were introduced to math functions (more in class 2)
- Sort data with `arrange()` and `arrange(desc())`
- Create summary statistics with `summarize()`
  
**Up next:** Watch videos for lab 2: vectors and data types. Try the warm-up

## Question from Piazza

- @38 Debugging with your eyes.
- @41 `tidyverse` vs `baseR`

## Practice

How many children live in Ohio?

- Restrict data to Ohio only.
- We have `poptotal` and `popadults`. Create a new column called `pop_children`.
- Aggregate data from Ohio counties to Ohio state total.

```{r, eval = FALSE}
# midwest is a data set that comes with the tidyverse
library(tidyverse)
midwest
```




## Feedback: 

```{r, echo = FALSE}
feedback_data  %>% 
  count(exp, pace) %>%
  ggplot(aes(y = n, x = pace, fill = exp)) + 
  geom_col() +
  labs(fill = "Starting\ncoding", x = "")
```


## Response:

- Adding 30 minutes before lab for more direct instruction. 
  - This is geared towards students who feel we're moving too fast. 

- Now that we know you better, TAs will group more deliberately.

- We will provide more "basics" practice.
  - "Optional" worksheets that should be doable after watching the videos
  
- We will also provide code from the pdf, so when you need to copy and paste you can use an R file.
  

## Course logistics:

- When should we start working on the final project?
    - Start looking for a dataset now. 
    - Write code to read it into R and start investigating with `dplyr` verbs.
      - Ask simple questions that can be addressed with your current tools.



# Vectors
## What's the deal with fall coding lab?

Two tracks:

**Accelerated**: 2 lessons

  - 2 lessons covering loops and functions.

**Not accelerated**: 5 lessons. 

  - 3 lessons review summer camp material 
  - 2 lessons covering loops and functions
  
**Logistics for both tracks**  
- Instructors Ari and Terence + wonderful TAs
- 80 minutes per week
- Not graded but will have a final project
- Access to TAs for coding specific problems throughout the quarter
  
## Key points: Class 2 vectors and data types

**vectors and vectorized coding**

- Vectors are the fundemental way to store data in R
- We can operate on vectors element-by-element without loops
  - `dplyr` verbs rely on this!
- We introduced built-in functions to build vectors and do operations on vectors. 

**data types**

- (Atomic) Vectors have a single data type
  - most often: `logical`, `integer`, `double`, or `character`
- Certain operations expect a certain data type and will try to coerce the data if it can.
  - coercion can lead to unexpected behavior such as making `NA`s.
  
**Up next:** Watch videos for lab 3: using `ifelse` for control flow.

## Exercise

Use R to calculate the sum 

$$\sum_{n=0}^{10} \frac{1}{2^n} = \frac{1}{2^0} + \frac{1}{2^{1}} + ..+\frac{1}{2^{10}} $$
$$\sum_{n=0}^{10} \frac{1}{2^n} = 1 + 0.5 + ..+ 0.00098 $$


1. Use vectorized math to create a vector with the correct numbers
1. Use a built-in function to add up all the numbers in the vector.

**Bonus** What happens to the sum as you increase `n`?

## Automatic type coercion

Type coercion is done automatically when R knows how. Usually, simpler types can be coerced to more complex types. 

  - `logical < integer < double < character`.

```{r}
# paste0() is a function that combines two chr into one
paste0("str", "ing")
paste0(1L, "ing")
```

`1L` is an `int`, but R will coerce it into a `chr` in this context.

## Automatic coercion

Logicals are coercible to numeric or character. This is very useful!

Determine the rule for how R treats `TRUE` and `FALSE` in math.

```{r, eval=FALSE}
TRUE + 4
FALSE + 4
sum(c(FALSE, FALSE, FALSE, FALSE))
mean(c(TRUE, TRUE, FALSE, FALSE, TRUE))
```

## Automatic coercion

```{r, eval=TRUE}
TRUE + 4
FALSE + 4
sum(c(FALSE, FALSE, FALSE, FALSE))
mean(c(TRUE, TRUE, FALSE, FALSE, TRUE))
```



## Question from Piazza Type coercion

@63

```{r, eval = FALSE}
1/n

Error in 1/n : non-numeric argument to binary operator


1/as.numeric(n)

Error in as.numeric(n) : 
  cannot coerce type 'closure' to vector of type 'double'
```

# control flow

## Key points: control flow with `if` and contingent column creation with `ifelse`


- Use `ifelse()` with `mutate()` to create new columns contingently.
  
    - `ifelse()` is vectorized so can operate on a logical vector to produce new results

- Understand how logical operators (i.e. `!`, `|`, `&`) work together with `ifelse` and conditional operators.

- Use `if()` (and `else`) to control whether an action is completed outside of a data context. 


We also introduced `Rmd`s and saw how to knit the `Rmd` to html or pdf.

**Up next:** watch video for class 4 on using `group_by` to do grouped analysis.^[This is way more exciting than it sounds.] Read the introduction to lab 4 before lab tomorrow!





## Execercise

We want to make a new column called `famous_storm` that is `1` for "Katrina" and "Rita" and `0` otherwise.


This code fails.

```{r, eval = FALSE}
# bad code :(
storms %>%
  mutate(famous_storm = 
           ifelse(name == "Katrina" | "Rita", 1, 0))
```

Find two alternative ways to write this code that work.


## Example: Creating a simulation dataset

You want to understand the impact of discrimination on gifted education.

```{r}
discrimination_simulation <- 
  tibble(group = rep(c(1, 2), 5000),
         probabilty_gets_tested = runif(10000),
         iq = rnorm(10000))
```

- Students in group 1 get tested with probability 60 percent
- Students in group 2 get tested with probability 10 percent
- Students get gifted education if `iq > 1` and they're tested


## Question from piazza

@87

What are the order of operations when using vectorized functions within `mutate()`?

```{r, eval = FALSE}

practice <- tibble(col = c(1, NA, 1, NA, 4))

practice %>%
  mutate(col = 
          ifelse(is.na(col), 
                 var(col, na.rm = TRUE), 
                 col)
         )

```


## Question from piazza

@82 (@79)

"How to aggregate across rows?"

```{r}
df <- tibble(sam = c(1, 0, 1, 1), 
             casey = c(1, 1, 0, 0), 
             abdul = c(1, 0, NA, 0))

df %>%
  mutate(rank = sum(sam, casey, abdul, na.rm = TRUE))
```



## Discussion questions

**warm-up take aways**

- `NA | TRUE` returns `TRUE`. Why does `FALSE | NA` return `NA`?
- `TRUE & NA` returns `NA`. Why does `FALSE & NA` return `FALSE`?

# Grouped analysis

## Key points: Grouped analysis with `group_by()`


- *groups* are a set of rows that belong together.

  - `group_by()` adds information about groups without changing the "data"

- Use `group_by()` with `summarize()` to create summary tables at group-level.

  - Use with functions that *reduce* data from a vector to a single value per group.
  
  - Expected output: a table with one row per group and one column per summary statistic and one column per grouping column.
  
- we can also use `group_by` to do grouped analysis with 

  - `mutate` with window functions or to add a summary stat as column for further analysis.
  
  - it also can impact `arrange` and `filter`

**Up next:** watch video for class 5 on using `ggplot`. In lab 5, your TA will review `ggplot` and then you'll have time to get support on your final project.

## Exercise

Rewrite this code to

\footnotesize
```{r, eval = FALSE}
wid_data %>%
  filter(percentile == "p99p100")  %>%
  group_by(country) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE))

wid_data %>%
  filter(percentile == "p90p100")  %>%
  group_by(country) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE))
```


##  Discussion

How do we make the data for this graph?

- What "groups" are required for the visualization?

```{r, echo = FALSE, message = FALSE, fig.height = 3}
mean_share_per_country_with_time <-
  wid_data %>% 
  filter(country %in% c("China", "India", "USA")) %>%
  filter(percentile %in% c("p99p100", "p90p100")) %>% 
  mutate(time_period = case_when(year < 1960 ~ "1959 and earlier",
                                 year < 1980 ~ "1960 to 1979",
                                 year < 2000 ~ "1980 to 1999",
                                 TRUE ~ "2000 to present")) %>%
  group_by(country, percentile, time_period) %>% 
  summarise(mean_share = mean(value, na.rm = TRUE),
            sd_share = sd(value, na.rm = TRUE))


 mean_share_per_country_with_time %>%
      ggplot(aes(x = country, y = mean_share, fill = percentile)) +
        geom_col(position = "dodge2") + 
        facet_wrap(~time_period)
```



## Exercise

A student came during office hours and asked why `mean(percentile=="p90p100")` doesn't calculate the average wealth shares (`value`) for the top 10 percentile group. 

```{r}
wid_data %>%
  filter(country %in% c("China", "India", "USA")) %>%
  group_by(country) %>%
  summarize(p90_mean = mean(percentile == "p90p100", 
                            na.rm = TRUE))
```

## Where does `n` come from?
\footnotesize
@98

```{r, eval = FALSE}
 disposition_by_race <-
    traffic_data %>% 
      mutate(Disposition = str_to_lower(Disposition),
             Disposition = 
               case_when(
                 Disposition %in% citation_strings ~ "citation",
                 Disposition %in% arrest_strings ~ "arrest",
                 TRUE ~ Disposition)) %>%
      count(Race, Disposition) %>%
      group_by(Race) %>%
      mutate(freq = round(n / sum(n), 3))  
```

