---
title: "Tutorial Read Me"
output: 
  html_notebook:
    toc: true
    toc_float: true
---

Hi all! This is an R notebook. This is where you'll be doing a lot of your
coding, especially when piloting code or when you want to present your 
results to others.

All the tutorials for Fall Coding Lab will consist of notebooks which will be
mostly pre-filled for you, and your goal will be to figure out the remaining
code. This serves several purposes: 

1. It familiarizes you with the file format you'll likely be interacting with 
the most while coding in R.
2. It familiarizes you with working within RStudio in general.
3. You don't have to write code from scratch, which can be daunting. 

Unfortunately, it does mean that because you're adhering to our "skeleton" for the 
code, the range of solutions you can come up with can sometimes be limited. So 
as you go through the code, if you think of a different solution, feel free to 
ignore the code we gave you and try it out! The worst that can happen is you have to 
undo your changes. Coding is *very* flexible like that.

# Some Basics

R notebooks are written in a language called R Markdown, so the scripts that 
generate them have the .Rmd file format. These are the files you'll be editing
in your lab assignments. They're very nice because you can write plain text,
like what you're reading right now, but you can also embed R code, run it, and
show the results as well:

```{r}
2 + 2
```

The way you tell RStudio that there's code to run is to insert a block of code,
called a chunk, starting with \`\`\`{r} and ending with \`\`\`. The keyboard 
shortcut to insert one is Ctrl + Alt + I on Windows, or Cmd + Option + I on OS X. 
There are many options for these chunks, that you can read up on here:

https://rmarkdown.rstudio.com/lesson-3.html

And on the topic of keyboard shortcuts, there are lots of them! You can find a
cheatsheet here:

https://rstudio.com/wp-content/uploads/2016/01/rstudio-IDE-cheatsheet.pdf

And a complete listing here:

https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts

No need to memorize them all, but if there's something you do often, it's 
better to learn the shortcut than to mouse around the user interface.

Once your code is ready to go, you can press Ctrl + Enter (Win) or Cmd + Return 
(OS X) to run the current line, or Ctrl + Shift + Enter or Cmd + Shift + Return
to run the entire current chunk. Try both out!

```{r}
print("COVID stinks.")
print("But at least I know some R Markdown now.")
```

Another command you'll likely use is Ctrl + Alt + R or Cmd + Option + R, which
runs the entire file.

# An Coding Lab Example

So now that you have an idea of how to work with a notebook, here's an example
of how a coding lab tutorial might look like.

Assume we're trying to write some simple code to solve the following scenario:
in non-COVID times, you're having a meal at Nella, Medici, or Shinju with 3 
friends, and you get a $75 bill. You want to figure out tax, tip, and the split.
What do you do?

The code you'll receive could look like:

```{r, eval = FALSE}
bill <- 75
tax <- 0.1075
tip <- # Tip appropriately!
number_people <- 

# Calculate total bill    
total <- bill + bill * (tax + tip)

# Split total equally between everyone
split <- 
```

And all you have to do is fill it out and run it:

```{r}
bill <- 75
tax <- 0.1075
tip <- 0.20 # Tip appropriately!
number_people <- 4

# Calculate total bill    
total <- bill + bill * (tax + tip)

# Split total equally between everyone
split <- total/number_people
```

So how much do you owe?

```{r}
split
```

Obviously the examples we'll use will be more complicated, but you get the 
gist of it. Some time after the actual session, we'll release the completed
notebooks so you can check your answers. 

# Text Formatting

R Markdown is a very complete language, and there are a lot of formatting 
options for text. You have all the usual suspects, **bold**, *italics*, etc.
If you're familiar with plain Markdown, great, they're pretty much the same. 
If not, here's a cheatsheet you can refer to:

https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

# Output Formats

R Markdown is very flexible in terms of output formats, but for this lab we'll 
stick to HTML notebooks. In this same folder, you'll find an identically named 
file ending in `nb.html`. This is a notebook that holds all of this text, code,
and results in an html file you can open in any browser, or share with others.
Every time you save your .Rmd, you're also saving your notebook.

However, this is only scratching the surface. You can make PDFs and many other
things using the same R Markdown script. Take a look at some examples at this
gallery:

https://rmarkdown.rstudio.com/gallery.html

If you get a chance, look up how to generate PDFs from an R Markdown file, this
will be super helpful for some of your classes where you have to turn in neatly
formatted empirical analyses!

