---
title: "The basics: 02 Vectors and data types"
author: "Ari Anisfeld"
date: "9/8/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Questions

## Create vectors

 1. In the lecture, we covered `c()`, `:`, `rep()`, `seq()` among other ways to create vectors.
 
    ```{r}
    dolly <- c(9, 10, 11, 12, 13, 14, 15, 16, 17)
    bees <- c("b", "b", "b", "b", "b")
    ```
 
  - Recreate `dolly` using `:`.
  - Create the same vector using `seq()`.
  - Recreate `bees` using `rep()`.

## Random vectors

1. In the lecture, we also created vectors using `rnorm()` and `runif()`. 

    ```{r}
    random_norm <- rnorm(100)
    random_unif <- runif(1000)
    ```
    
    - How long are the vectors `random_norm` and `random_unif`? Use `length()` to verify.
    - What are the largest and smallest values in `random_norm` and `random_unif`? Use `min()` and `max()`.
    - Use `mean()` and `sd()` to calculate the mean and standard deviation of the two distributions.
    - Create a new vector with `10000` draws from the standard normal distribution. 
    - `rnorm()` by default sets `mean = 0` (see `?rnorm`). Create a vector of `10000` draws from the normal distribution with `mean = 1`. Use `mean()` to verify.
    
    
Notice the functions `min()`, `max()`, `mean()` and `sd()` all take a vector with many values and summarize them as one value. These are good to use with `summarize()` when doing data analysis on tibbles.

## data types

- Use `typeof()` to verify the data types of `dolly`, `bees`, `random_unif`

- Coerce `dolly` to a character vector. Recall we have functions `as.<type>()` for this kind of coercion.

- Try to coerce `bees` to type `numeric`. What does R do when you ask it to turn "b" into a number?

## vectorized math

`a` and `b` are vectors of length 10. Look at them in the console.

```{r}
a <- 1:10
b <- rep(c(2, 4), 5)
```


1. Add `a` and `b` element by element.
1. Subtract `a` and `b` element by element.
1. Divide `a` by `b` element by element.
1. Multiply `a` and `b` element by element.
1. Raise the element of `a` to the power of `b` element by element.
1. Multiply each element of `a` by 3 then subtract `b`
1. Raise each element of `b` to the third power.
1. Take the square root of each element of `a`.

## vectorized comparison

1. Run the following code and make sure you understand the output.

```{r}
a > b
a == b
```

## creating tibbles with vectors

1. Create a tibble with columns called `a` and `b` where `a` is the numbers 1 to 100 and `b` is 100 random numbers from the standard normal distribution.

```{r, eval = FALSE}
my_tibble<- 
tibble(
  a = ...,
  b = ...
)
```


## subsetting

`midwest` is a data set that comes with `ggplot`.

```{r, eval = FALSE}
# try this to see midwest data
library(tidyverse)
midwest %>% head()
```

1. Use `pull()` to get the vector of `state` names.
1. Use `[[` to get a vector of `state` names. This is baseR and requires (normal) quotes  around column names.
1. Use `select()` to get a tibble with `state` as the only column.
1. Use `[` to get a tibble with `state` as the only column


# Solutions

## Create vectors

```{r}
dolly_colon <- 9:17
dolly_seq <- seq(9:17)
bees_rep <- rep("b", 5)
```

## Random vectors

```{r}
# lengths
length(random_norm)
length(random_unif)

# largest and smallest values (repeat with random_unif)
max(random_norm)
min(random_norm)

# mean and sd  (repeat with random_unif)
mean(random_norm)
sd(random_norm)

# rnorm with length 10000
longer_rnorm <- rnorm(10000)

# mean = 1
rnorm_centered_on_one <- rnorm(10000, mean = 1)

```


## typeof

```{r}
typeof(dolly)
typeof(bees)
typeof(random_unif)

# notice dolly is int if it's created by : or seq
typeof(dolly_seq)
typeof(dolly_colon)

# coercion
as.character(dolly)

# R coerces "b" to NA because there is 
# not a natural number to replace "b" with
as.numeric(bees)
```
## vectorized math

```{r}
a + b
a - b
a / b
a * b
a ^ b
2 * a - b
b ^ 3
sqrt(a)
```


## Creating tibbles

```{r}
my_tibble<- 
tibble(
  a = seq(1, 100),
  b = rnorm(100)
)
```

```{r}
midwest %>% pull(state)
midwest[["state"]]

midwest %>% select(state)
midwest[,"state"]
```

