---
title: "The basics: 03 ifelse"
author: "Ari Anisfeld"
date: "9/8/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Questions

## `ifelse`

We'll start using `ifelse` which is commonly used in data analysis with `mutate()`.


`midwest` is a dataset built into `tidyverse`

1. create a new variable called `poverty_designation` that is "High Poverty" if `percbelowpoverty` is above 10 and is "Low Poverty" otherwise.

    If you pipe your tibble into `count(poverty_designation)`, you should see
    
    ```{r, echo = FALSE}
    midwest %>%
      mutate(poverty_designation = ifelse(percbelowpoverty > 10, "High Poverty", "Low Poverty")) %>%
      count(poverty_designation)
      
    ```

1. Create a new variable that is "Ohio Counties" for observations from Ohio and "Other Midwestern Counties" for the rest of the observations.

1. Create a new variable that is `TRUE` for the observations from the counties "COOK", "WAYNE", "CUYAHOGA", "OAKLAND" or "FRANKLIN"  and `FALSE` otherwise. Use the `%in%` operator.

1. In this problem, we'll simulate an election.

```{r}
election_simulation <- 
  tibble(probabilty_vote = runif(1000),
         probability_support = runif(1000))
```

    a. Using `mutate` and `ifelse` create a new column called `voter` that is 1 if the `probablity_vote` is over .5 and 0 otherwise.
    
    b. Create a second column called `supporter` that is 1 if `probablity_support` is over .4 and 0 otherwise.
    
    c. Create a third column that equals `TRUE` if `voter` and `supporter` are both equal to 1, that equals `FALSE` if `voter` equals 1 but `supporter` is 0 and that is `NA` otherwise.^[There are several ways forward, e.g. using `case_when` or nesting two `ifelse()` statements.]


## Using `if`

We use `if()` when working on "statistical programming" (ie. when not working with tibbles for data analysis). We'll develop a small dice game. 

1. Fill in the ... so the code says "You win" if the dice add up to 7 and "You lose" otherwise.

    ```{r, eval = FALSE}
    dice <- sample(c(1:6), 2)
    
    if (...) {
      print("You win")
    } else {
      print("You lose")
    }
    
    ```



1. Add an `else if()` block to the code above that says try again if the dice add up to 6 or 8.




## Solution 
1.
    ```{r, eval = FALSE}
    midwest %>%
      mutate(poverty_designation = ifelse(percbelowpoverty > 10, "High Poverty", "Low Poverty")) %>%
      count(poverty_designation)
      
    ```

1. 
    ```{r, eval = FALSE}
    midwest %>%
      mutate(ohio = ifelse(state == "OH", "Ohio Counties", "Other Midwestern Counties"))
    ```

1.
    ```{r, eval = FALSE}
    big_counties <- c("COOK", "WAYNE", "CUYAHOGA", "OAKLAND", "FRANKLIN")
    
    midwest %>%
      mutate(populous_counties = ifelse(county %in% big_counties, 1, 0)
    ```


1. 
    ```{r}
    simulation <- 
      tibble(probabilty_vote = runif(1000),
             probability_support = runif(1000)) %>%
      mutate(voter = ifelse(probabilty_vote > .5, 1, 0),
             supporter = ifelse(probability_support > .4, 1, 0),
             results = case_when(voter == 1 & supporter == 1 ~ TRUE,
                                 voter == 1 & supporter == 0 ~ FALSE,
                                 TRUE ~ NA))
    
    # An alternative approach takes advantage of the structure of the data
    
    simulation <- 
      tibble(probabilty_vote = runif(1000),
             probability_support = runif(1000)) %>%
      mutate(voter = ifelse(probabilty_vote > .5, 1, 0),
             supporter = ifelse(probability_support > .4, 1, 0),
             results = ifelse(voter == 1, supporter * voter, NA))
              
    ```
    
    
## if

    ```{r}
    dice <- sample(c(1:6), 2)
    
    if (sum(dice) == 7) {
      print("You win")
    } else if(sum(dice) %in% c(6,8)) {
      print ("Try again")
    } else {
      print("You lose")
    }
    
    ```