---
title: "The basics: 04 grouped analysis"
author: "Ari Anisfeld"
date: "9/8/2020"
urlcolor: blue
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform = FALSE)
library(tidyverse)
```

# Questions


## `group_by` and `summarize`

1. `midwest` is a data set that comes bundled with `tidyverse`. In an earlier lab you calculated the population of Ohio in the following way.

    ```{r, eval = FALSE}
    midwest %>%
      filter(state == "OH")
      summarize(total_population = sum(poptotal))
    ```
    
    With `group_by` you can calculate the total population of all the states at once!
    
    ```{r, eval= FALSE}
    midwest %>%
      group_by(...) %>%
      summarize(total_population = sum(poptotal))
    ```

1. For each state in the `midwest` data, calculate total `area`.

1. For each state in the `midwest` data, calculate the proportion of counties that are in a metro area (`inmetro`).^[Recall that the `mean()` of a column of `0` and `1`s tell you the proportion of `1`s.]

1. For each state, calculate the proportion of people with a college degree and also with high school degrees.
  
    - First, use `mutate` to calculate the number of people with the degree type.
    - Then, use `group_by` and `summarize` to calculate the proportions. 

## `group_by` and `mutate`

1. Add a column to midwest called `pop_state` that equals the state population. Compare your result to what you calculated early. 

    ```{r, eval = FALSE}
    # fill in the ... with approriate code
    midwest %>%
      group_by( ... ) %>%
      mutate(pop_state = ... )
    ```

1. Building off the previous question, create a column that shows the number of people living below the poverty line (`percbelowpoverty`) in each county. Look at your results to make sure they make sense.

## `count`

1. Reproduce this table using `count()`.
    ```{r, echo = FALSE}
    midwest %>%
      group_by(inmetro) %>%
      summarize(n = n())
    ```

1. Reproduce this table using `add_count()`.
    ```{r, echo = FALSE}
    midwest %>%
      select(state, inmetro) %>%
      group_by(inmetro) %>%
      mutate(n = n()) %>%
      head()
    ```
```{r, eval = FALSE}
# fill in the ... with the appropriate code.
 midwest %>%
    select(state, inmetro) %>%
    ... %>%
    head()

```

1. Reproduce the following table

    ```{r, echo = FALSE}
    midwest %>%
      count(state, inmetro)
    ```


Want to improve this tutorial? Report any suggestions/bugs/improvements on [here](mailto:anisfeld@uchicago.edu)! We’re interested in learning from you how we can make this tutorial better.

# Solutions

##
1.  
    ```{r, eval= FALSE}
    midwest %>%
      group_by(state) %>%
      summarize(total_population = sum(poptotal))
    ```
1. 

    ```{r, eval= FALSE}
    midwest %>%
      group_by(state) %>%
      summarize(total_area = sum(area))
    ```

1.
    ```{r, eval = FALSE}
    midwest %>%
      group_by(state) %>%
      summarize(prop_in_metro = mean(inmetro))
    
    ```




1.
    ```{r, eval = FALSE}
     midwest %>%
      mutate(pop_with_hs = perchsd * poptotal,
             pop_with_college = percollege * poptotal) %>%
      group_by(state) %>%
      summarize(total_population = sum(poptotal),
                perc_with_hs = sum(pop_with_hs)/total_population,
                perc_with_college = sum(pop_with_college)/total_population,)

    ```
      You might have been tempted to do it in the following way, but this underestimates the statewide rates. Why?

    ```{r, eval = FALSE}
      midwest %>%
        group_by(state) %>%
        summarise(perc_with_hs = mean(perchsd))
    ```

## `group_by` and `mutate`

1.

    ```{r, eval = FALSE}
    midwest %>%
      group_by(state) %>%
      mutate(pop_state = sum(poptotal)) 
    ```

1. A careful analyst would say this is wrong, because we do not know the poverty status of each and every person in the counties (see `percpovertyknown`). A challenge problem is to find the lower and upper bound on the number of people with poverty per county.

    ```{r, eval = FALSE}
    midwest %>%
      group_by(state) %>%
      mutate(pop_state = sum(poptotal),
             pop_below_poverty = pop_state * percbelowpoverty/100)
    ```

## `count`

1. 
    ```{r}
    midwest %>%
      count(inmetro)
    ```

1. 
    ```{r, eval = FALSE}
    # fill in the ... with the appropriate code.
     midwest %>%
        select(state, inmetro) %>%
        add_count(inmetro) %>%
        head()
    
    ```
    
1.

    ```{r, echo = FALSE}
    midwest %>%
      count(state, inmetro)
    ```