---
title: "The basics: 06 functions"
author: "Ari Anisfeld"
date: "9/8/2020"
urlcolor: blue
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform = FALSE)
library(tidyverse)
```

# Questions


## Writing functions

Recall a function has the following form

```{r, eval = FALSE}
name <- function(args) {
  # body
  do something (probably with args)
}
```


1. Write a function called `calc_quadratic` that takes an input `x` and calculates $f(x) = x^2 + 2x + 1$. For example:

    ```{r, echo = FALSE}
    calc_quadratic <- function(x) {
      x ^ 2 + 2 * x + 1
    }
    ```
    
    ```{r}
    calc_quadratic(5)
    ```
  
    a. What are the arguments to your function? What is the body of the function?
    a. This function is vectorized! (Since binary operators are vectorized). Show this is true by running `calc_quadratic` with an input vector that is -10 to 10. 

1. You realize you want to be able to work with any quadratic. Update your functions so that it can work with any quadratic in standard form $f(x) = ax^2 + bx + c$. 

    - Your new function will take arguments `x`, `a`, `b` and `c`. 
    - Set the default arguments to `a=1`, `b=2` and `c=1`
  
  
1. Write a function called `solve_quadratic` that takes arguments `a`, `b` and `c` and provides the two roots using the [quadratic formula](https://en.wikipedia.org/wiki/Quadratic_formula).
  
    In our outline, we suggest you:
    - Calculate the determinant ($\sqrt{b^2 - 4ac}$) and store as an intermediate value. 
    - Return two values by putting them in a vector. If you stored the roots as `root_1` and `root_2`, then the final line of code in the function should be `c(root_1, root_2)` or, if you prefer, `return(c(root_1, root_2))`.
      
    
    ```{r, eval = FALSE}
    # fill in the ... with appropriate code
    solve_quadratic <- function(...){
      
      determinant <- ...
      root_1 <- ...
      root_2 <- ...
      
      c(root_1, root_2)
    
      }
    ```
    
    ```{r, echo = FALSE}
    solve_quadratic <- function(a, b, c){
      
      determinant <- sqrt(b ^ 2 - 4 * a * c)
      root_1 <- (-b + determinant) / (2 * a)
      root_2 <- (-b - determinant) / (2 * a)
      
      c(root_1, root_2)
    
      }
    ```
    
    The code should work as follows:
    
    ```{r}
    solve_quadratic(a = -4, b = 0, c = 1)
    ```

1. We "normalize" a variable by subtracting the mean and dividing by the standard deviation $\frac{x - \mu}{\sigma}$. Write a function called `normalize` that takes a vector as input and normalizes it.

    
    ```{r, echo = FALSE}
    
    # Note using the name `na.rm` is good practice, because it is
    # consistent with other R functions and so less likely to be frustrating
    normalize <- function(x, na.rm = FALSE) {
      (x - mean(x, na.rm = na.rm)) / sd(x, na.rm = na.rm)
    }
    ```

    You should get the following output.
    ```{r}
    normalize(1:5)
    ```

    a. What output do you get when the input vector is `0:4`? How about `-100:-96`? Why?
    a. What happens when your input vector is c(1,2,3,4,5, NA)? Rewrite the function so the result is:^[Hint: take advantage of `mean` and  `sd` NA handling.]
    
        ```{r, echo = FALSE}
        normalize(c(1,2,3,4,5, NA), na.rm = TRUE)
        ```
    
    a. The `txhousing` data set is comes with `ggplot`. Use your normalize function in `mutate` to create  `normalized_annual_volume` to make the following graph.

    ```{r, eval = FALSE}
    # replace the ... with the appropriate code.
    txhousing %>%
      group_by(year, city) %>%
      summarize(annual_volume = sum(volume, na.rm = TRUE)) %>%
      group_by(year) %>%
      mutate(...) %>%
      ggplot(aes(x = year, y = normalized_annual_volume)) + 
      geom_point() +
      geom_line(aes(color = city))
    ```
    ```{r, echo = FALSE, message = FALSE}
    txhousing %>%
      group_by(year, city) %>%
      summarize(annual_volume = sum(volume, na.rm = TRUE)) %>%
      group_by(year) %>%
      mutate(normalized_annual_volume = normalize(annual_volume)) %>%
      ggplot(aes(x = year, y = normalized_annual_volume)) + 
      geom_point() +
      geom_line(aes(color = city))
    ```

Want to improve this tutorial? Report any suggestions/bugs/improvements on [here](mailto:anisfeld@uchicago.edu)! We’re interested in learning from you how we can make this tutorial better.


# Solutions

## Writing functions

1. Write a function called `calc_quadratic` that takes an input `x` and calculates $f(x) = x^2 + 2x + 1$. For example:

    ```{r, echo = TRUE}
    calc_quadratic <- function(x) {
      x ^ 2 + 2 * x + 1
    }
    ```
    
    a. What are the arguments to your function? What is the body of the function?
      
      **arguments are `x`; the body is `x ^ 2 + 2 * x + 1`**
    
    a. This function is vectorized! (Since binary operators are vectorized). Show this is true by running `calc_quadratic` with an input vector that is -10 to 10. 
    
    
    ```{r}
    calc_quadratic(-10:10)
    ```

1. You realize you want to be able to work with any quadratic. Update your functions so that it can work with any quadratic in standard form $f(x) = ax^2 + bx + c$. 

    - Your new function will take arguments `x`, `a`, `b` and `c`. 
    - Set the default arguments to `a = 1`, `b = 2` and `c = 1`
    
    ```{r, echo = TRUE}
    calc_quadratic <- function(x, a = 1, b = 2, c = 1) {
     a * x ^ 2 + b * x + c
    }
    
    calc_quadratic(5)
    ```
  
  
1. Write a function called `solve_quadratic` that takes arguments `a`, `b` and `c` and provides the two roots using the [quadratic formula](https://en.wikipedia.org/wiki/Quadratic_formula).
  
    
    ```{r, echo = TRUE}
    solve_quadratic <- function(a, b, c){
      
      determinant <- sqrt(b ^ 2 - 4 * a * c)
      root_1 <- (-b + determinant) / (2 * a)
      root_2 <- (-b - determinant) / (2 * a)
      
      c(root_1, root_2)
    
      }
    ```
    
    The code should work as follows:
    
    ```{r}
    solve_quadratic(a = -4, b = 0, c = 1)
    ```
    
    Notice, the code doesn't deal with functions with no roots. It returns `NaN`. If there is a single root (such as when a = 1, b = 0 and c = 0), it returns the same number twice. We could use `if()` statments in the function to have it explicitly deal with these issues.

1. We "normalize" a variable by subtracting the mean and dividing by the standard deviation $\frac{x - \mu}{\sigma}$. Write a function called `normalize` that takes a vector as input and normalizes it.

    
    ```{r, echo = FALSE}
    
    # Note using the name `na.rm` is good practice, because it is
    # consistent with other R functions and so less likely to be frustrating
    normalize <- function(x, na.rm = FALSE) {
      (x - mean(x, na.rm = na.rm)) / sd(x, na.rm = na.rm)
    }
    ```

    You should get the following output.
    ```{r}
    normalize(1:5)
    ```

    a. What output do you get when the input vector is `0:4`? How about `-100:-96`? Why?
    
      **You get the same results as `1:5`. This is because when you demean all the vectors are identical.**
    
    a. What happens when your input vector is c(1,2,3,4,5, NA)? Rewrite the function so the result is:^[Hint: take advantage of `mean` and  `sd` NA handling.]
    
      **see above**
    
    a. The `txhousing` data set is comes with `ggplot`. Use your normalize function in `mutate` to create  `normalized_annual_volume` to make the following graph.

    ```{r, echo = TRUE, eval = FALSE}
    txhousing %>%
      group_by(year, city) %>%
      summarize(annual_volume = sum(volume, na.rm = TRUE)) %>%
      group_by(year) %>%
      mutate(normalized_annual_volume = normalize(annual_volume)) %>%
      ggplot(aes(x = year, y = normalized_annual_volume)) + 
      geom_point() +
      geom_line(aes(color = city))
    ```
