---
title: "fall_2021_meetings"
author: "Ari Anisfeld"
date: "9/1/2021"
output: 
  beamer_presentation:
    fig_width: 6
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
#knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

# Class 1: Reading files and `dplyr'


## Do now 
**Do now:**

  - Complete the intro poll at `bit.ly/acc_intro_poll`
  - Let us know if you do not have R and RStudio installed!
  
  
\footnotesize
**After the poll**

  - Download `fall_lab_1` from the course webpage: `harris-coding-lab.github.io`.


\normalsize

## Expectations

From you:

  - do the work
  - engage in course!
  - have R and RStudio installed!

From us:

  - prepare engaging lesson materials to teach you R basics
  - address your questions about R
  - help you gain self-confidence and self-sufficiency

From everyone:

  - be nice to each other and create a growth-focused environment

<!-- Ask them to think of anything else -->


## Do the work 

  - Step 1. Videos
  - Step 1a. Basics
  - Step 2. Live sessions
  - Step 3. Finish Labs

## R basics

We cover:

  - how to read in and manipulate data
  - how to work with basic data structures
    - tibbles
    - vectors
  - programmer logic 
    - if statements
    - loops
    - functions

## R we don't cover 

**Your homework! That's your professors' gift to you!**

We won't cover **in depth**:

  - most statistical tools
  - advanced tools like webscrapping, package development and so forth
  - some fundemental data skills
    - how to join data together 
    - how to convert data from long to wide (pivoting)
    - how to deal with very messy data
    - how to work with specific data types (e.g. dates, advanced strings)
  

## Address your questions about R

  - Ed discussions (linked on canvas)
  - Office hours (.....)
  - Here!
  
  *Tip: If you're stuck on your homework because of R, reframe your question as an R question.*

## And help you build self-sufficency



## And help you build self-sufficency

  - use ?
  - test code in console. try to break it.
  - ask teammates / try googling
  - ask us! (Ed / Office hours / live now)

If it's not "mission critical", you can safely move on without full understanding. 
(Imagine learning a language and trying to figure out all the grammar and vocabulary at the same time!)
<!-- go over column types live --- read about col_types --> 

## In this sessions
  
  Walk away confident in reading in data  and manipulating data with `dplyr` verbs.
  
  - Warm-up
  - Guided practice -- setting up 
  - Independent or group practice
  - comprehension check
  
  

## Warm up 

1. How would you install the `haven` package? 


## Warm up 

2. In the videos, you learned about `head()`. What if you wanted to get the tail end of your data instead?


## Warm up

3. Recall our `dplyr` verbs. 


What is the purpose of each function?

- mutate()
- filter()
- select()
- arrange()
- summarize()

Soon we'll add:

- group_by() 



## Warm up

4. Imagine you have a data set, `df` with 4 variables, `county`, `year`, `income`, and `employment`. You only need the year and employment status of people whose income is below $5000. Which two `dplyr` commands do you need to do this? Can you write the code for this?

5. Remember the `mean()` function? What `dplyr` commands would we need if we want the average income? How many rows will the resulting dataset be?
 
 
## Question: Can you explain pipes?

- Pipes `%>%` take the left hand side and put them into the first position on the right hand side.

```{r, eval = FALSE}
storms %>% filter(year > 2010) %>% glimpse()

recent_storms <- filter(storms, year > 2010)
glimpse(recent_storms)
```
Notice

- `filter()` takes data in the first position and then an arbitrary number of filtering expressions.
- `glimpse()` takes data in the first position

 
## Quick look at `group_by()`

Compare:

```{r}
txhousing %>%
  summarize(total_volume = sum(volume, na.rm=TRUE))

txhousing %>%
  group_by(year) %>%
  summarize(total_volume = sum(volume, na.rm=TRUE))
```


...




## Lesson 0: Intro to R, RStudio and the `tidyverse`

- navigate and use Rstudio's features
  - particularly, the console, the text editor and help
- assign objects to names with `<-`
- use functions by providing inputs and learn more with `?`
- `install.packages()` (once) and then load them with `library()` (each time you restart R)

## Lesson 1: Key points: Reading files

- Tabular data is stored in a lot of different formats.
  - e.g. `.csv`, `.xlsx`, `.dta`
- Read tabular data of a given type with the proper function.
  - e.g. for `csv`s we have `read_csv()`
  - If you get a new type, Google "How to read xxx files into R tidyverse".
- We need to be aware of the file path and can `setwd()`.
- We know there are useful tools built into the `read_xxx()` functions.
  - Though we just scratched the surface.
  
## Lesson 1: Manipulating data with `dplyr()`

- Choose columns with `select()`.
- Choose rows based on a match criteria with `filter()`.
  - We were introduced to comparison operators like  `==` and `%in%`.
- Make new columns with `mutate()`.
- Sort data with `arrange()` and `arrange(desc())` or `arrange(-x)`.
- Create summary statistics with `summarize()`.





  