---
title: "summer_2021_qa"
author: "Ari Anisfeld"
date: "9/1/2021"
output: 
  beamer_presentation:
    fig_width: 6
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
#knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

# Class 1: Reading files and `dplyr'


## Do now 
**Do now:**

  - Complete the intro poll at `bit.ly/acc_intro_poll`
  
  
\footnotesize
**After the poll**

  - Download `lab_1` from the course webpage: `harris-coding-lab.github.io`.
  
**Notice**  

  - Earlier we had a typo for the link to `lab_0` on canvas, which is now fixed. Sorry for inconvience.  

\normalsize

## Expectations

From you:

  - do the work (i.e. watch video, try basics, do lab, bring questions)
  - engage in course! (i.e. work with partners, answer questions, do polls)
  - have R and RStudio installed!

From us:

  - prepare engaging lesson materials
  - address your questions
  - help you be confident for core (confident != R expert)
  
From everyone:

  - be nice to each other and create a growth-focused environment

<!-- Ask them to think of anything else -->


## Do the work 

  - Step 1. Videos
  - Step 1a. Basics
  - Step 2. QA
  - Step 3. Lab 

## Not an expert

We cover:
  
  - how to work with basic data structures (tibbles, vectors)
  - how to read and manipulate data
  - programmer logic (if statements, loops, functions)

We won't cover in depth:

  - most statistical tools
  - how to join data together 
  - how to convert data from long to wide (pivoting)
  - how to deal with very messy data
  - how to work with specific data types (e.g. dates, advanced strings)
  - among other things like webscrapping, package development and so forth
    
    
## Today's session

  - Set up working directories
  - Review some questions from QA
  - Highlight key points and open up for live questions
       
## Setting up working directory and coding environment
  
  1. Do you have a folder on your computer for coding lab material?  
  If not, create one and make sure you know the path to the folder.

  2. We recommend creating a `problem_set` folder inside your coding lab folder.
  
  3. Make folder called `data` inside the `problem_set` folder. 
  

## Putting your files in place

  4. Create a new R script. Save your script in the `problem_set` folder. From now on, when you start a script or `Rmd` save it there.

  5. Download the first data set [here](https://github.com/harris-coding-lab/harris-coding-lab.github.io/raw/master/data/world_wealth_inequality.xlsx) and put the data in your `data` folder. Find the link in the lab pdf!
  
## Tell R where to find files

  - Local paths are like addresses on your computer.  Use `getwd()` to see how your computer paths look.

\footnotesize

```{r, eval = FALSE}
# In lab0 we downloaded data form a URL which is an address on the internet. 
covid_data <- 
  read_csv(
    "https://data.cdc.gov/api/views/qfhf-uhaa/rows.csv?"
    ) 

# Compare to a local path
wid_data <- 
  read_xlsx(
    "~/coding-lab/harris-coding-lab.github.io/data/world_wealth_inequality.xlsx"
  )
```
\normalsize
  6. Add a line to your script where you `setwd()` to the `data` folder.

## Working with the files

  7. Finally, we are using data in an excel format. We need the package `readxl` to process data of this type. In the console, run `install.packages("readxl")`.

  8. Add code to load the `tidyverse`.    

  9. If you followed the set-up from above, you should be able to run the following code with no error.

```{r, eval = FALSE}
wid_data <- read_xlsx("world_wealth_inequality.xlsx")
```
  
  <!-- look at briefly and point out that \n is an issue 
      wid_data$`shweal_p99p100_z_CN
    Net personal wealth
    Top 1% | share
    `
  --> 
    
## What to do when something is confusing?

  - use ?
  - test code in console. try to break it.
  - ask teammates / try googling
  - ask us!

If it's not "mission critical", you can safely move on without full understanding. 
(Imagine learning a language and trying to figure out all the grammar and vocabulary at the same time!)
<!-- go over column types live --- read about col_types --> 

## Question: 

- What's the deal with `col_types = cols(Suppress = col_character())`?
- Do we need that "accessType=DOWNLOAD&bom=true&format=true%20target=" part?
\footnotesize
    ```{r, eval = FALSE}
    covid_data <- 
      read_csv(
        paste0("https://data.cdc.gov/api/views/qfhf-uhaa/rows.csv?",
                "accessType=DOWNLOAD&bom=true&format=true%20target="), 
        col_types = cols(Suppress = col_character()))
    ```
\normalsize
- Note: In URLs after the `?` you send meta information about your request. 


## Question: Can you explain pipes?

- Pipes `%>%` take the left hand side and put them into the first position on the right hand side.

```{r, eval = FALSE}
storms %>% filter(year > 2010) %>% glimpse()

recent_storms <- filter(storms, year > 2010)
glimpse(recent_storms)
```
Notice

- `filter()` takes data in the first position and then an arbitrary number of filtering expressions.
- `glimpse()` takes data in the first position


## Lesson 0: Intro to R, RStudio and the `tidyverse`

- navigate and use Rstudio's features
  - particularly, the console, the text editor and help
- assign objects to names with `<-`
- use functions by providing inputs and learn more with `?`
- `install.packages()` (once) and then load them with `library()` (each time you restart R)

## Lesson 1: Key points: Reading files

- Tabular data is stored in a lot of different formats.
  - e.g. `.csv`, `.xlsx`, `.dta`
- Read tabular data of a given type with the proper function.
  - e.g. for `csv`s we have `read_csv()`
  - If you get a new type, Google "How to read xxx files into R tidyverse".
- We need to be aware of the file path and can `setwd()`.
- We know there are useful tools built into the `read_xxx()` functions.
  - Though we just scratched the surface.
  
## Lesson 1: Manipulating data with `dplyr()`

- Choose columns with `select()`.
- Choose rows based on a match criteria with `filter()`.
  - We were introduced to comparison operators like  `==` and `%in%`.
- Make new columns with `mutate()`.
- Sort data with `arrange()` and `arrange(desc())` or `arrange(-x)`.
- Create summary statistics with `summarize()`.

  
# Class 2: Vectors and data types  

## Course logistics:

- When should we start working on the final project?
    - Start looking for a dataset now. 
    - Write code to read it into R and start investigating with `dplyr` verbs.
      - Ask simple questions that can be addressed with your current tools.

**lab 1 solutions** will be available on the course website.
 

## Getting started with Rmarkdown (Rmd)

- What's an Rmd?
- How to make an Rmd
- How to work with an Rmd 

## Knitting: making the frustrating part less frustrating

- Install `tinytex`

```{r, eval = FALSE}
  install.packages("tinytex")
  tinytex::install_tinytex()
```

- Knit early and often. 


## When to use Rmds vs scripts?

**Rmd**

- Exploration of data
- Presentations and reports

**script**

- Projects with interrelated code (e.g. an R package)
- Working on a server that does not have Rstudio installed


## Questions from QA

**Question 1**:
- Why do I need the function `summarize` in the following bit of code? 

```{r} 
michigan_population_total <- 
  midwest %>% 
    filter(state == "MI") %>% 
  summarize(total_pop = sum(poptotal))
```

- Why can't I just pipe directly into `sum`?

**Question 2**: Why do we need to use `pull()`?


## R's primary data structures: Vectors vs. tibbles

- Why do we need different data structures?
- How are vectors and tibbles related?


## Why do we need different data structures?

In theory, we can do all our work with vectors.

```{r}
names_vec <- c("Ari", "Qiwei", "Jay", "Thomas")
surnames_vec <- c("Anisfeld", "Lin",  "Zaleski", "Whamond")
position_vec <- c("Instructor", "TA", "TA", "TA")
```

why might I want a tibble?

## R's primary data structures: Vectors vs. tibbles

Tibbles encapsulate vectors

```{r}
names_vec <- c("Ari", "Qiwei", "Jay", "Thomas")
surnames_vec <- c("Anisfeld", "Lin",  "Zaleski", "Whamond")
positions_vec <- c("Instructor", "TA", "TA", "TA")

my_tibble <- tibble(names = names_vec,
                    surnames = surnames_vec,
                    positions = positions_vec)
```

- Keep data tidy -> rows are a single observation or record.
- Keep meta-data (column names)
- Keep track of relationships between vectors
- Can hold various data types

## R's primary data structures: Vectors vs. tibbles

- vectors are simpler
- have a single data type
- some functions expect vectors or make more sense on them.



## Reviewing automatic type coercion

Type coercion is done automatically when R knows how. Usually, simpler types can be coerced to more complex types. 

  - `logical < integer < double < character`.

```{r}
# paste0() is a function that combines 
# two chr vectors into a single vector
paste0("str", "ing")
paste0(1L, "ing")
```

`1L` is an `int`, but R will coerce it into a `chr` in this context.

## Automatic coercion

Logicals are coercible to numeric or character. This is very useful!

Determine the rule for how R treats `TRUE` and `FALSE` in math.

```{r, eval=FALSE}
TRUE + 4
FALSE + 4
sum(c(FALSE, FALSE, FALSE, FALSE))
mean(c(TRUE, TRUE, FALSE, FALSE, TRUE))
```

## Automatic coercion

```{r, eval=TRUE}
TRUE + 4
FALSE + 4
sum(c(FALSE, FALSE, FALSE, FALSE))
mean(c(TRUE, TRUE, FALSE, FALSE, TRUE))
```



## Exercise

**1**: Use R to calculate the sum 

$$\sum_{n=0}^{10} \frac{1}{2^n} = \frac{1}{2^0} + \frac{1}{2^{1}} + ..+\frac{1}{2^{10}} $$
$$\sum_{n=0}^{10} \frac{1}{2^n} = 1 + 0.5 + ..+ 0.00098 $$


1. Use vectorized math to create a vector with the correct numbers
1. Use a built-in function to add up all the numbers in the vector.

**Bonus** What happens to the sum as you increase `n`?


**2**: Use `paste0()` to convert `v1` and `v2` into `"hello!"`

```{r}
v1 <- c("h","l", "o")
v2 <- c("e", "l", " !")
```

```{r, echo = FALSE}
paste(v1,  v2, collapse = "")
```





## Key points: Class 2 vectors and data types

**vectors and vectorized coding**

- Vectors are the fundemental way to store data in R
- We can operate on vectors element-by-element without loops
  - `dplyr` verbs rely on this!
- We introduced built-in functions to build vectors and do operations on vectors. 

**data types**

- (Atomic) Vectors have a single data type
  - most often: `logical`, `integer`, `double`, or `character`
- Certain operations expect a certain data type and will try to coerce the data if it can.
  - coercion can lead to unexpected behavior such as making `NA`s.
  
**Over weekend:** Attempt lab 2. 
**For Tuesday:** Watch video about control flow + try basics.


# Class 3: Control flow

## Outline

- lingering questions about Rmds
- `ifelse()` questions

## Reviewing the anatomy of an Rmd:

*Write text in the document*

```{r} 
# ^^^ start an R chunk ```{r} 
# sometimes {} have meta information

# R code goes in a chunk 
ex <- seq(1, 12)

# and output prints below
print(ex)
```


## A meta examples: {r, echo = FALSE}

```{r, echo = FALSE}
print(ex)
```


## Another meta examples: {r, eval = FALSE}


```{r, eval = FALSE}
print(ex)
```

## Naming chunks

Chunk names help you debug

```{r example, echo = TRUE}
# code
```

Two chunks cannot have the same name

```{r example2}
# If I change the name to "example" we will get an error.
```


## Quick hitters

*Q:* Why don't we get LaTex with \$ in this example?

```{r, eval = FALSE}
$$
  \sqrt{p}
$$
```


## Quick hitters


*Q:* Why don't we get LaTex with \$ in this example?
*A:* In a code chunk, knitr expects R code! So a dollar sign is interpreted  to be referring to a named entity in an object
like `data$column_name`.  

So put LaTex in the "text" area of an Rmd!

$$
\sqrt{p}
$$


## Quick hitters

*Q:* What does this error tell you?

```{r, eval = FALSE}
setwd("~/Documents/coding")
```
*Error in setwd("~/Documents/coding") : cannot change working directory*

## Quick hitters

*Q:* What does this error tell you?
*A:* Usually, it means your directory name is wrong somehow!

```{r, eval = FALSE}
setwd("~/Documents/coding_lab/")
```

Note: "~/" does not work on Windows! 

## An aside on relative paths

You can refer to directories **relative** to your current directory using `.` and `..`

- `.` means current directory.
- `..` means "go back" the directory path.


## an example

File structure:

- coding_lab
  datafile.csv
  - problem_sets
    my_rmd.Rmd
    

You could access the data with `read_csv("../datafile.csv")` in a chunk.


## What's the difference between `&` and `&&`?

- Test your hypothesis with additional examples in the console?

```{r}
c(TRUE, TRUE) & c(TRUE, FALSE)
c(TRUE, TRUE) && c(TRUE, FALSE)
```

- Which plays nicely with `ifelse()`?
- Which plays nicely with `if()`?

## What's the difference between `|` and `||`?

Similarly OR has a vectorized `|` and singleton `||` version.

- generally, you can get by with `|` and `&`! 
- when you are working on code for general use (like writing a package) there will be times when you want to ensure 

## Exercise

We want to make a new column called `famous_storm` that is `1` for "Katrina" and "Rita" and `0` otherwise.

This code fails.

```{r, eval = FALSE}
# bad code :(
storms %>%
  mutate(famous_storm = 
           ifelse(name == "Katrina"| "Rita", 1, 0))


storms %>%
  mutate(famous_storm = 
           ifelse(name == "Katrina"| name == "Rita", 1, 0))


storms %>%
  mutate(famous_storm = 
           ifelse(name %in% c("Katrina", "Rita", "Amy", "Bobby"), 1, 0))
```

Find two alternative ways to write this code that work.


## Example: Creating a simulation dataset

You want to understand the impact of discrimination on gifted education.

```{r}
discrimination_simulation <- 
  tibble(group = rep(c(1, 2), 5000),
         prob_tested = runif(10000),
         iq = rnorm(10000)) 
```

- Students in group 1 get tested with probability 60 percent
- Students in group 2 get tested with probability 10 percent
- Students get gifted education if `iq > 1` and they're tested




## Key points: control flow with `if` and contingent column creation with `ifelse`


- Use `ifelse()` with `mutate()` to create new columns contingently.
  
    - `ifelse()` is vectorized so can operate on a logical vector to produce new results

- Understand how logical operators (i.e. `!`, `|`, `&`) work together with `ifelse` and conditional operators.

- Use `if()` (and `else`) to control whether an action is completed outside of a data context. 


We also introduced `Rmd`s and saw how to knit the `Rmd` to html or pdf.

**Up next:** prepare lab 3 for tomorrow. **Friday** watch video for class 4 on using `group_by` to do grouped analysis.


##  Key points:`if()` versus `ifelse()`

|| `if()`/ `else()` | `ifelse()` |
|---- | ---- | ---- |
| Used to conditionally evaluate code | yes | yes|
| Vectorized? | no, only uses first element | yes|
| Handles NA | no, error `missing value where TRUE/FALSE needed` | yes, returns `NA` |
| baseR | yes | yes^[there's a tidyverse `if_else()` that works slightly differently] |


- *Takeaway:* When we are focusing on data analysis use `ifelse()` (or `if_else()`).



## Quick hitters

**Class 2 basics**

- *Q* Why did I get double when your code shows the type is an int?

```{r}
typeof(seq(1, 12, 1))
```


## Quick hitters

**Class 2 basics**

- *Q* Why did I get double when your code shows the type is an int?
- *A* In base R, data types are not always predictable. 

```{r}
typeof(seq(1, 12))
typeof(seq(1, 12, 1))
```


Tidyverse functions tend to be more careful to avoid this sort of behavior.


## Question from QA, what's the deal with subsetting?

How many ways can you pull out the even numbers from `vec`?

```{r}
vec <- 1:10
vec[c(2, 4, 6, 8, 10)]
vec[rep(c(FALSE, TRUE), 5)]
vec[seq(2, 10, 2)]
vec[vec %% 2 == 0]
```

Notice `filter()` works like the final option!



## Horoscope game:

Make a game where you ask the user to enter their birth month and you tell them their fortune. 

- Give people born in December, January or February a "cold" fortune
- Give people born in June through September a "warm" fortune
- Give people born in November a great fortune
- Give everyone else an okay fortune

e.g.  `birth_month <- 2` the code should return something like `I see penguins in your forecast`.




# Class 4: Grouped analysis

## practice

We are working in `our_rmd.Rmd` with file structure like so:

- dir1
  file1.csv
  - dir2
    our_rmd.Rmd
  - dir3
    file2.csv

**Question**

- What is our working directory?
- How do we access file2.csv?
- How do we access file1.csv?

# Class 5: Loops

# Class 6: Functions
